### 2024-03-28 paper/paper.qmd
removed wild turkey data and updated preambles

---
title: "Wildlife Hunting..."
author: Samantha Barfoot
thanks: "Code and data are available at: LINK."
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---
# Introduction

# Data

# Model

# Results
 - Averages, means, etc (summary statistics)
 - Differences over time

# Discussion

# References

### 2024-03-28 scripts/00-simulate_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
-# Purpose: Simulates... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Purpose: Simulates... 
+# Author: Samantha Barfoot
+# Date: 28 March 2024
+# Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# Any other information needed? 
 
 
 #### Workspace setup ####

### 2024-03-28 scripts/01-download_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
 # Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Author: Samantha Barfoot
+# Date: 28 March 2024
+# Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# Any other information needed?
 
 
 #### Workspace setup ####

### 2024-03-28 scripts/02-data_cleaning.R

@@ -1,11 +1,11 @@
 #### Preamble ####
 # Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 6 April 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Author: Samantha Barfoot
+# Date: 28 March 2024
+# Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# Any other information needed?
 
 #### Workspace setup ####
 library(tidyverse)

### 2024-03-28 scripts/03-test_data.R

@@ -1,11 +1,11 @@
 #### Preamble ####
 # Purpose: Tests... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Author: Samantha Barfoot
+# Date: 28 March 2024
+# Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# Any other information needed?
 
 
 #### Workspace setup ####

### 2024-03-28 scripts/04-model.R

@@ -1,11 +1,11 @@
 #### Preamble ####
 # Purpose: Models... [...UPDATE THIS...]
-# Author: Rohan Alexander [...UPDATE THIS...]
-# Date: 11 February 2023 [...UPDATE THIS...]
-# Contact: rohan.alexander@utoronto.ca [...UPDATE THIS...]
+# Author: Samantha Barfoot
+# Date: 28 March 2024
+# Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: [...UPDATE THIS...]
-# Any other information needed? [...UPDATE THIS...]
+# Pre-requisites: 
+# Any other information needed?
 
 
 #### Workspace setup ####

### 2024-04-03 paper/paper.qmd
Simulated data for bears, deer, wolves, and moose

@@
 ---
-title: "Wildlife Hunting..."
+title: "Analysis of Ontario Wildlife Hunting Data"
+subtitle: "2009-2023"
 author: Samantha Barfoot
@@
 
+The analysis, graphs, and tables were made with [@citeR].
+
 # Data
@@
 # Results
- - Averages, means, etc (summary statistics)
- - Differences over time
 
@@
 
+## What was done
+
+## Something we learned about the world
+
+## Another thing we learned about the world
+
+## Weaknesses
+
+## Looking forward
+
 # References

### 2024-04-03 scripts/00-simulate_data.R

@@ -1,19 +1,175 @@
 #### Preamble ####
-# Purpose: Simulates... 
+# Purpose: Simulates the moose, black bear, white tailed deer, wolf and coyote harvest data from the Ontario Data Catalogue
 # Author: Samantha Barfoot
 # Date: 28 March 2024
 # Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
 # Pre-requisites: 
-# Any other information needed? 
-
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(janitor)
 
 #### Simulate data ####
-# [...ADD CODE HERE...]
+set.seed(1)
 
+#variable that represents the number of columns in the simulated data set
+num_samples <- 1700
 
+# White tailed deer simulated data #
+#simulated tibble table
+simulated_data_deer <-
+  tibble(
+    # Randomly pick an option, with replacement, 153 times
+    "WMU" = sample(
+      x = 0:100,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Year" = sample(
+      x = 2009:2023,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Active Hunters" = sample(
+      x = 0:500,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Antlered Harvests" = sample(
+      x = 0:200,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Antlerless Harvests" = sample(
+      x = 0:200,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Total Harvests" = sample(
+      x = 0:400,
+      size = num_samples,
+      replace = TRUE
+    )
+  )
 
+# Wolf and coyote tailed deer simulated data #
+#simulated tibble table
+simulated_data_wolf_coy <-
+  tibble(
+    # Randomly pick an option, with replacement, 153 times
+    "WMU" = sample(
+      x = 0:100,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Year" = sample(
+      x = 2009:2023,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Active Hunters" = sample(
+      x = 0:500,
+      size = num_samples,
+      replace = TRUE
+    ),
+    "Harvest" = sample(
+      x = 0:200,
+      size = num_samples,
+      replace = TRUE
+    )
+  )
+
+# Moose simulated data #
+#simulated tibble table
[…]

### 2024-04-04 paper/paper.qmd
Updated quarto heading and cleaned moose data

@@
 ---
-title: "Analysis of Ontario Wildlife Hunting Data"
+title: "Analysis of Ontario Wildlife Harvests and Active Hunters"
 subtitle: "2009-2023"

### 2024-04-04 scripts/00-simulate_data.R

@@ -16,122 +16,13 @@ set.seed(1)
 #variable that represents the number of columns in the simulated data set
 num_samples <- 1700
 
-# White tailed deer simulated data #
-#simulated tibble table
-simulated_data_deer <-
-  tibble(
-    # Randomly pick an option, with replacement, 153 times
-    "WMU" = sample(
-      x = 0:100,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Year" = sample(
-      x = 2009:2023,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Active Hunters" = sample(
-      x = 0:500,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Antlered Harvests" = sample(
-      x = 0:200,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Antlerless Harvests" = sample(
-      x = 0:200,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Total Harvests" = sample(
-      x = 0:400,
-      size = num_samples,
-      replace = TRUE
-    )
-  )
-
-# Wolf and coyote tailed deer simulated data #
-#simulated tibble table
-simulated_data_wolf_coy <-
-  tibble(
-    # Randomly pick an option, with replacement, 153 times
-    "WMU" = sample(
-      x = 0:100,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Year" = sample(
-      x = 2009:2023,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Active Hunters" = sample(
-      x = 0:500,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Harvest" = sample(
-      x = 0:200,
-      size = num_samples,
-      replace = TRUE
-    )
-  )
-
-# Moose simulated data #
-#simulated tibble table
-simulated_data_moose <-
-  tibble(
-    # Randomly pick an option, with replacement, 153 times
-    "WMU" = sample(
-      x = 0:100,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Year" = sample(
-      x = 2009:2023,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Active Hunters" = sample(
-      x = 0:500,
-      size = num_samples,
-      replace = TRUE
-    ),
-    "Bull Harvests" = sample(
-      x = 0:200,
-      size = num_samples,
-      replace = TRUE
[…]

### 2024-04-04 scripts/01-download_data.R

@@ -9,18 +9,11 @@
 
 
 #### Workspace setup ####
-library(opendatatoronto)
 library(tidyverse)
-# [...UPDATE THIS...]
+library(stringr)
 
 #### Download data ####
-# [...ADD CODE HERE TO DOWNLOAD...]
 
 
 
 #### Save data ####
-# [...UPDATE THIS...]
-# change the_raw_data to whatever name you assigned when you downloaded it.
-write_csv(the_raw_data, "inputs/data/raw_data.csv") 
-
-         

### 2024-04-04 scripts/02-data_cleaning.R

@@ -9,36 +9,40 @@
 
 #### Workspace setup ####
 library(tidyverse)
+library(MASS)
+library(stringr)
 
-#### Clean data ####
-raw_data <- read_csv("inputs/data/plane_data.csv")
+#### Read in data ####
+raw_data_moose <- read_csv("data/raw_data/moose_2023.csv",
+                          show_col_types = FALSE,
+                          skip = 1)
 
-cleaned_data <-
-  raw_data |>
-  janitor::clean_names() |>
-  select(wing_width_mm, wing_length_mm, flying_time_sec_first_timer) |>
-  filter(wing_width_mm != "caw") |>
-  mutate(
-    flying_time_sec_first_timer = if_else(flying_time_sec_first_timer == "1,35",
-                                   "1.35",
-                                   flying_time_sec_first_timer)
-  ) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "490",
-                                 "49",
-                                 wing_width_mm)) |>
-  mutate(wing_width_mm = if_else(wing_width_mm == "6",
-                                 "60",
-                                 wing_width_mm)) |>
-  mutate(
-    wing_width_mm = as.numeric(wing_width_mm),
-    wing_length_mm = as.numeric(wing_length_mm),
-    flying_time_sec_first_timer = as.numeric(flying_time_sec_first_timer)
-  ) |>
-  rename(flying_time = flying_time_sec_first_timer,
-         width = wing_width_mm,
-         length = wing_length_mm
-         ) |> 
-  tidyr::drop_na()
+#raw_data_bear <- read_csv("data/raw_data/black_bear_2023.csv",
+#                          show_col_types = FALSE,
+#                          skip = 1)
+
+#### Clean in data ####
+num_records_moose <- dim(raw_data_moose)[1]
+
+#creates matrix full of zeros to store clean data
+moose_data <- matrix(rep(0, 7*36), ncol = 7, nrow = 36)
+
+#fills clean data matrix with the cleaned data
+#columns are as follows: wildlife management region, year, active hunters, bull harvest,
+# cow harvest, calf harvest, and total harvest
+for(i in 1:num_records_moose){
+  #unit <- raw_data_moose[[i,1]]
+  unit <- "075"
+  if(str_starts(unit,"0")){
+    str_remove(unit,"0")
+  }
+  if(str_ends(unit,"[A-Za-z]")){
[…]

### 2024-04-04 paper/paper.qmd
Updated abstract

--- 2024-04-03 paper/paper.qmd
@@
 ---
-title: "Analysis of Ontario Wildlife Hunting Data"
+title: "Analysis of Ontario Wildlife Harvests and Active Hunters"
 subtitle: "2009-2023"
 author: Samantha Barfoot
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/wildlife_hunting.git"
 date: today
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper analyzes Ontario wildlife harvest counts, specifically moose and black bears, and the number fo active hunters based on region.
+The analysis examines harvest counts from 2012 - 2023 across northern, southeastern, and southwestern Ontario.
+It was found that..."
 format: pdf

### 2024-04-04 paper/paper.qmd
Updated title and subtitle of paper

@@
 ---
-title: "Analysis of Ontario Wildlife Harvests and Active Hunters"
-subtitle: "2009-2023"
+title: "Is it Better to be a Moose or a Black Bear?"
+subtitle: "An Analysis of Ontario Wildlife Harvest and Active Hunter Rates From 2012-2023"
 author: Samantha Barfoot

### 2024-04-15 paper/paper.qmd
Completed data cleaning script for moose and bear data

--- 2024-04-04 paper/paper.qmd
@@
 ---
-title: "Analysis of Ontario Wildlife Harvests and Active Hunters"
-subtitle: "2009-2023"
+title: "Is it Better to be a Moose or a Black Bear?"
+subtitle: "An Analysis of Ontario Wildlife Harvest and Active Hunter Rates From 2012 to 2023"
 author: Samantha Barfoot
-thanks: "Code and data are available at: LINK."
+thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/wildlife_hunting.git"
 date: today
 date-format: long
-abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
+abstract: "This paper analyzes Ontario wildlife harvest counts, specifically moose and black bears, and the number fo active hunters based on region.
+The analysis examines harvest counts from 2012 to 2023 across northern, southeastern, and southwestern Ontario.
+It was found that..."
 format: pdf
 number-sections: true
+table-of-contents: true
 bibliography: references.bib

### 2024-04-15 scripts/01-download_data.R

@@ -14,6 +14,11 @@ library(stringr)
 
 #### Download data ####
 
-
+unit <- "75A"
+if(str_ends(unit,"[A-Za-z]")){
+  unit <- str_remove(unit,"[A-Za-z]")
+}
+unit <- strtoi(unit)
+print(unit)
 
 #### Save data ####

### 2024-04-15 scripts/02-data_cleaning.R

@@ -11,38 +11,133 @@
 library(tidyverse)
 library(MASS)
 library(stringr)
+library(dplyr)
 
 #### Read in data ####
 raw_data_moose <- read_csv("data/raw_data/moose_2023.csv",
-                          show_col_types = FALSE,
-                          skip = 1)
+                          show_col_types = FALSE)
 
-#raw_data_bear <- read_csv("data/raw_data/black_bear_2023.csv",
-#                          show_col_types = FALSE,
-#                          skip = 1)
+raw_data_bear <- read_csv("data/raw_data/black_bear_2023.csv",
+                          show_col_types = FALSE)
 
 #### Clean in data ####
 num_records_moose <- dim(raw_data_moose)[1]
+num_records_bear <- dim(raw_data_bear)[1]
 
-#creates matrix full of zeros to store clean data
-moose_data <- matrix(rep(0, 7*36), ncol = 7, nrow = 36)
-
-#fills clean data matrix with the cleaned data
-#columns are as follows: wildlife management region, year, active hunters, bull harvest,
-# cow harvest, calf harvest, and total harvest
+#cleans WMU names so that there are no letters or beginning zeros for moose
 for(i in 1:num_records_moose){
-  #unit <- raw_data_moose[[i,1]]
-  unit <- "075"
+  unit <- raw_data_moose[[i,1]]
+  #Checks if the wmu value has a zero at the start or a letter at the end of the string and removes them
   if(str_starts(unit,"0")){
-    str_remove(unit,"0")
+    unit <- str_remove(unit,"0")
   }
   if(str_ends(unit,"[A-Za-z]")){
-    str_remove(unit,"[A-Za-z]")
+    unit <- str_remove(unit,"[A-Za-z]")
   }
-  unit <- strtoi(unit)
-  print(unit)
+  raw_data_moose[[i,1]] <- unit
+}
+
+#cleans WMU names so that there are no letters or beginning zeros  for bears
+for(i in 1:num_records_bear){
+  unit <- raw_data_bear[[i,1]]
+  #Checks if the wmu value has a zero at the start or a letter at the end of the string and removes them
+  if(str_starts(unit,"0")){
+    unit <- str_remove(unit,"0")
+  }
+  if(str_ends(unit,"[A-Za-z]")){
[…]

### 2024-04-16 paper/paper.qmd
Created graphs for data section in quarto document

@@
 ---
+
+```{r}
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
+
+library(tidyverse)
+library(ggplot2)
+```
+
 # Introduction
@@
 # Data
+```{r}
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
+#### Read in data ####
+north_bear <-
+  read_csv(
+    file = "../data/analysis_data/north_data_bear.csv",
+    show_col_types = FALSE
+  )
+se_bear <-
+  read_csv(
+    file = "../data/analysis_data/southeast_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+sw_bear <-
+  read_csv(
+    file = "../data/analysis_data/southwest_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+north_moose <-
+  read_csv(
+    file = "../data/analysis_data/north_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+se_moose <-
+  read_csv(
+    file = "../data/analysis_data/southeast_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+sw_moose <-
+  read_csv(
+    file = "../data/analysis_data/southwest_data_moose.csv",
+    show_col_types = FALSE
+  )
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-north
+#| fig-cap: "Northern harvest numbers for moose and bears between 2009 and 2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-southeast
+#| fig-cap: "Southeastern harvest numbers for moose and bears between 2009 and 2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-southwest
+#| fig-cap: "Southwestern harvest numbers for moose and bears between 2009 and 2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-bearharvest
+#| fig-cap: "Bear harvest numbers for the northern, southeastern and southwestern regions from 2012-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "red")+
+  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "green")+
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-mooseharvest
+#| fig-cap: "Moose harvest numbers for the northern, southeastern and southwestern regions from 2012-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "blue") + 
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "green")+
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-swhunters
+#| fig-cap: "Number of active bear and moose hunters in the southwest region from 2009-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
+  geom_line(data = sw_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-sehunters
+#| fig-cap: "Number of active bear and moose hunters in the southeast region from 2009-2012"
+
+#create line graph 
+ggplot()+
+  geom_line(data = se_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
+  geom_line(data = se_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-northhunters
+#| fig-cap: "Number of active bear and moose hunters in the northern region from 2009-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = north_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
+  geom_line(data = north_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-swharvesthunters
+#| fig-cap: "Number of active bear hunters plotted against total bear harvests from 2012-2023 in the southwest region"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-seharvesthuntersbear
+#| fig-cap: "Number of active bear hunters plotted against total bear harvests from 2012-2023 in the southeast region"
+
+#create line graph 
+ggplot()+
+  geom_line(data = se_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-northharvesthuntersbear
+#| fig-cap: "Number of active bear hunters plotted against total bear harvests from 2012-2023 in the northern region"
+
+#create line graph 
+ggplot()+
+  geom_line(data = north_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-swharvesthuntersmoose
+#| fig-cap: "Number of active moose hunters plotted against total moose harvests from 2009-2023 in the southwest region"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-seharvesthuntersmoose
+#| fig-cap: "Number of active moose hunters plotted against total moose harvests from 2009-2023 in the southeast region"
+
+#create line graph 
+ggplot()+
+  geom_line(data = se_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-northharvesthuntersmoose
+#| fig-cap: "Number of active moose hunters plotted against total moose harvests from 2009-2023 in the northern region"
+
+#create line graph 
+ggplot()+
+  geom_line(data = north_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  theme_minimal()
+```
 

### 2024-04-20 paper/paper.qmd
Completed all graphs and wrote the data and results sections of the Quarto document

--- 2024-04-15 paper/paper.qmd
@@
 title: "Is it Better to be a Moose or a Black Bear?"
-subtitle: "An Analysis of Ontario Wildlife Harvest and Active Hunter Rates From 2012 to 2023"
+subtitle: "An Analysis of Ontario Wildlife Harvest and Active Hunter Data From 2006 to 2023"
 author: Samantha Barfoot
@@
 date-format: long
-abstract: "This paper analyzes Ontario wildlife harvest counts, specifically moose and black bears, and the number fo active hunters based on region.
+abstract: "This paper analyzes Ontario wildlife harvest counts, specifically moose and black bears, and the number of active hunters based on region.
 The analysis examines harvest counts from 2012 to 2023 across northern, southeastern, and southwestern Ontario.
@@
 ---
+
+```{r}
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
+
+library(tidyverse)
+library(ggplot2)
+library(gridExtra)
+```
+
 # Introduction
 
-The analysis, graphs, and tables were made with [@citeR].
+Every year the Ontario Ministry of Natural Resources and Forestry records data on the number of black bears and moose hunted per year and the number of active hunters per year.
+Ontario moose hunting season typically runs from early September to late October [@moose] and black bear season runs from early May to mid June or mid August to late November depending on the region [@bear].
+To hunt either of these animals, hunters must have the proper licence and are responsible for identifying their target and obtaining tags.
+Moose tags indicate what type of moose a hunter is allowed to harvest whether it be a cow, calf, or bull [@moose].
+Hunting in Ontario is a significant contributor to the Ontario economy and tourism industry.
+It supports the employment of thousands of people and some years generates more money than the film and television industry in Ontario [@huntAs].
+One year Canadian hunters spent around \$1.2 billion on hunting trips with \$70 on the hunting licences alone [@huntAs].
+In 2000, the hunting industry represented \$1.5 billion of the economy [@huntAs].
+Despite being such a large industry there is still a significant amount of poaching and other illegal hunting practices performed in Ontario [@poachNum].
+In 2016 there were 384 major offences related to these practices [@poachNum].
 
-# Data
+Therefore, it is important to properly regulate and monitor the wildlife populations as well as the hunting and harvest numbers so that species are not put at risk by over hunting.
+Additionally, it is important to analyze how these levels change over the years and what might cause these changes.
+Alongside over hunting, having too large a population of a certain species can also put a strain on the ecosystem in which case hunting is an important method of regulating population.
+This paper explores the trends of active hunter and harvest numbers from 2006-2023 for moose and 2021-2023 for black bears.
 
-# Model
+WHAT WAS FOUND
 
-# Results
+The estimand of this paper is the correlation between the number of active hunters in Ontario and the number of annual moose and black bear harvests.
+Specifically, we examine how these numbers vary with regard to region and time
 
-# Discussion
+@sec-data of this paper covers the broader context of the Ontario moose and black bear hunting activty and harvest data sets from the Ontario Data Catalogue.
+It examines the variables used in the data set using graphs to help explain them as well as discussing high-level aspects of the data cleaning process.
+@sec-results further examines the relationship between the variables and the finding from the analysis of the two data sets.
+The @sec-discussion section of this paper explores what was learned from this analysis and the implications it has on Ontario's hunting industry and environment.
+Additionally, it identifies possible biases and factors that may have effected the analysis.
+
+MORE ABOUT THE SECTIONS
+
+The graphs and tables in this paper were made with [@citeR] using R studio.
+The creation of these graphs and tables were made with `ggplot` [@ggplotRef], `tiddyverse` [@tidyverseRef], `gridExtra` [@gridRef] packages.
+The analysis and cleaning of the data was conducted with and `dyplr` [@dplRef ], `MASS` [@massRef], `janitor` [@janitorRef] and `stringr` [@stringRef] packages.
+
+# Data {#sec-data}
+```{r}
+#| echo: false
+#| include: false
+#| warning: false
+#| message: false
+
+#### Read in data ####
+north_bear <-
+  read_csv(
+    file = "../data/analysis_data/north_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+se_bear <-
+  read_csv(
+    file = "../data/analysis_data/southeast_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+sw_bear <-
+  read_csv(
+    file = "../data/analysis_data/southwest_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+north_moose <-
+  read_csv(
+    file = "../data/analysis_data/north_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+se_moose <-
+  read_csv(
+    file = "../data/analysis_data/southeast_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+sw_moose <-
+  read_csv(
+    file = "../data/analysis_data/southwest_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+all_moose <-
+  read_csv(
+    file = "../data/analysis_data/all_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+types_moose <-
+  read_csv(
+    file = "../data/analysis_data/types_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+all_bear <-
+  read_csv(
+    file = "../data/analysis_data/all_data_bear.csv",
+    show_col_types = FALSE
+  )
+```
+
+## Data Source
+
+The data sets used in this report, *Moose hunting activity and harvests* and *Black bear hunting activity and harvests*, were obtained from the Ontario Data Catalogue.
+Specifically, the data is provided by the ministry of Natural Resources and Forestry and were last updated on February 16, 2024.
+They data sets provide the estimated hunter and harvest numbers by region across Ontario from 2006 for the moose data set and 2012 for the black bear data set.
+The numbers in these data sets are based on replies that the Ontario government received from a sample of hunters across Ontario.
+It is not specified exactly how these measurements were performed only that they received replies from these various hunters.
+The number of active hunters and harvests could have been measured by surveys or the government asking all hunters registered for moose and black bear hunting how many harvests they collected.
+To legally hunt moose and black bears, hunters must register and obtain licences.
+Therefore, to measure how many active hunters are in Ontario the Ministry of Resources and Forestry might have counted how many registered hunters they have for each species.
+However, they can only estimate how many harvests were collected each year as it depends on hunters being honest about how many harvests they made and that all hunters posses the proper permits.
+
+## Variables of Interest
+
+The original data set for moose hunting and harvests obtained from the Ontario Data Catalogue contained seven variables.
+The first variable,*WMU*, is the Wildlife Management Unit which indicates the specific region in Ontario.
+There are a total of 95 different WMUs which are grouped into three larger regions, northern, southeastern, and southwestern Ontario.
+The second variable, *Year*, indicates the harvest year.
+*Active Hunter* (see @fig-huntersmoose) indicates how many hunters were active in that specific WMU and year.
+*Bull Harvest*, *Cow Harvest*, and *Calf Harvest* (see @fig-harvestmoosetypes) indicate the respective number of harvests for each type of moose for that specific WMU and year.
+*Total Harvest* (see @fig-harvestmoose) is the sum of all bull, cow, and calf harvests for that specific WMU and year.
+The original data set for black bear hunting and harvests contains the same variables of WMU, year, and active hunters (see @fig-huntersbear).
+However, instead of bull, cow, calf, and total harvest there is only one harvest variable called *Harvest* (see @fig-harvestbear) as all black bears are grouped as one.
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-huntersmoose
+#| fig-cap: "Number of active moose hunters between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+
+plot1 <- 
+  ggplot()+
+  geom_line(data = north_moose, aes(x = Year, y = `Active Hunters`), color = "black") +
+  theme_minimal()+
+  labs(tag = "A")
+
+plot2 <-
+  ggplot()+
+  geom_line(data = se_moose, aes(x = Year, y = `Active Hunters`), color = "black") +
+  theme_minimal() +
+  labs(tag = "B")
+
+plot3 <-
+  ggplot()+
+  geom_line(data = sw_moose, aes(x = Year, y = `Active Hunters`), color = "black") +
+  theme_minimal() +
+  labs(tag = "C")
+
+grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2)
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-harvestmoosetypes
+#| fig-cap: "Number of cow, bull, and calf moose types harvested between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+
+plot1 <- 
+  ggplot()+
+  geom_line(data = types_moose, aes(x = Year, y = `Cow Harvest`), color = "black") +
+  theme_minimal()+
+  labs(tag = "A")
+
+plot2 <-
+  ggplot()+
+  geom_line(data = types_moose, aes(x = Year, y = `Calf Harvest`), color = "black") +
+  theme_minimal() +
+  labs(tag = "B")
+
+plot3 <-
+  ggplot()+
+  geom_line(data = types_moose, aes(x = Year, y = `Bull Harvest`), color = "black") +
+  theme_minimal() +
+  labs(tag = "C")
+
+grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2)
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-harvestmoose
+#| fig-cap: "Number of moose harvested between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+
+plot1 <- 
+  ggplot()+
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "black") +
+  theme_minimal()+
+  labs(tag = "A")
+
+plot2 <-
+  ggplot()+
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "black") +
+  theme_minimal() +
+  labs(tag = "B")
+
+plot3 <-
+  ggplot()+
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "black") +
+  theme_minimal() +
+  labs(tag = "C")
+
+grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2)
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-huntersbear
+#| fig-cap: "Number of active black bear hunters between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+
+plot1 <- 
+  ggplot()+
+  geom_line(data = north_bear, aes(x = Year, y = `Active Hunters`), color = "black") +
+  theme_minimal()+
+  labs(tag = "A")
+
+plot2 <-
+  ggplot()+
+  geom_line(data = se_bear, aes(x = Year, y = `Active Hunters`), color = "black") +
+  theme_minimal() +
+  labs(tag = "B")
+
+plot3 <-
+  ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = `Active Hunters`), color = "black") +
+  theme_minimal() +
+  labs(tag = "C")
+
+grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2)
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-harvestbear
+#| fig-cap: "Number of bears harvested between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+
+plot1 <- 
+  ggplot()+
+  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "black") +
+  theme_minimal()+
+  labs(tag = "A")
+
+plot2 <-
+  ggplot()+
+  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "black") +
+  theme_minimal() +
+  labs(tag = "B")
+
+plot3 <-
+  ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "black") +
+  theme_minimal() +
+  labs(tag = "C")
+
+grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2)
+```
+
+To clean the data, each animals data sets were split into three sets, one for WMUs in the northern, southeastern, and southwestern region.
+This way, different regions and their harvest and active hunter numbers could be plotted against each other.
+Each cleaned black bear data set contains 12 entries and each cleaned moose data set contains 18 entries.
+For each larger region, the active hunter and harvest numbers were totaled for each year.
+Each cleaned data set also contains the same variables as the original data sets.
+Some cleaning was also applied to the WMU values as many of them had extra zeros and letter joined to them.
+
+TABLE?
+
+# Results {#sec-results}
+
+## Total Harvests
+
+By analyzing each hunting region, northern, southeastern, and southwestern, separately, we can examine how the total harvest rates of black bears and moose compare.
+@fig-north shows that the harvest rates for moose generally decreases before 2018 whereas the amount of black bear harvests increases from 2012 to 2016 before decreasing and increasing again in 2021.
+The trend lines on @fig-north cross three times between 2012 and 2023 with black bear harvest rates sometimes being higher than moose rates and vice versa.
+In comparison, they never cross in @fig-southeast with the number of black bear harvests always being more than moose harvests.
+Interestingly, the peak for both @fig-north and @fig-southeast for black bears are both in 2016 and have similar trends where they dramatically increase before 2016 and quickly fall after 2016.
+Whereas, in @fig-southwest the black bear harvest peak stays relatively consistent between 2016 and 2019 before greatly decreasing after 2016.
+In @fig-southwest, similar to @fig-southeast the number of black bear harvests is always higher than the number of moose harvests except in 2020 where they are equal.
+Interestingly, in all three graphs, the number of moose harvests tend to decrease overall between 2006 and 2023, most dramatically in the northern and southwestern regions.
+While all three of these graphs show that black bear harvest numbers are generally higher than moose, the number of harvests for each species are drastically different.
+In the northern region moose harvests reach around 5500 and black bears around 5000.
+Whereas, in the southeast region they reach 1500 and 3000, respectively, and in the southwestern region, 350 and 275 respectively.
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-north
+#| fig-cap: "Northern harvest numbers for moose and bears between 2006 and 2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-southeast
+#| fig-cap: "Southeastern harvest numbers for moose and bears between 2006 and 2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-southwest
+#| fig-cap: "Southwestern harvest numbers for moose and bears between 2006 and 2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  theme_minimal()
+
+```
+
+We can also compare the harvest rates of each species separately for each region.
+@fig-bearharvest we can see more clearly what we observed previously where the amount of black bears harvested is very different between each region.
+In the northern region, on average, has the most the most harvests followed by the southeast and then southwest region which is significantly lower than the other two regions.
+However, the northern and southeastern rates cross over at two points around 2020.
+@fig-mooseharvest shows a similar trend with the northern region always having the most harvests followed by the southeastern then the southwestern regions with no intersections.
+Both species likely experience this difference in harvest numbers because of the regions size.
+The northern region is significantly larger than the southeast and southwest regions.
+However, it is interesting that the northern and southeastern harvest rates are so similar between 2018 and 2023 given that the areas are of drastically different size.
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-bearharvest
+#| fig-cap: "Bear harvest numbers for the northern, southeastern and southwestern regions from 2012-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "blue") + 
+  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "red")+
+  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "green")+
+  theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-mooseharvest
+#| fig-cap: "Moose harvest numbers for the northern, southeastern and southwestern regions from 2006-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "blue") + 
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "green")+
+  theme_minimal()
+```
+
+## Active Hunters
+
+Additionally, we can compare the number of active moose and black bear hunters in all three regions.
+From @fig-northhunters we can see that there are significantly more moose hunters than black bear hunters every year.
+However, the number of moose hunters has significantly decreased since 2006.
+In 2006 there was over 65,000 moose hunters and in 2023 there were only around 25,000 moose hunters.
+In comparison, the number of black bear hunters has remained relatively consistent compared to moose hunters since 2012.
+The southwest region (see @fig-swhunters) has a similar trend with the number of moose hunters greatly decreasing since 2006.
+Additionally, there has generally been less bear hunters in the region compared to moose hunters up until 2021 at which point the number of black bear hunters surpassed the number of moose hunters.
+The southeast region (see @fig-sehunters), however, has a different trend; the number of moose hunters in the region stayed relatively consistent from 2006 up to 2019 after which it dramatically decreased.
+Around this time the number of black bear hunters also surpassed the number of moose hunters.
+Since 2012 the number of black bear hunters has increased.
+The discrepancy between the number of hunters by regain could, again, be due to the fact that the southwestern and southeastern regions are much smaller than the northern region.
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-swhunters
+#| fig-cap: "Number of active bear and moose hunters in the southwest region from 2006-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = sw_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
+  geom_line(data = sw_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-sehunters
+#| fig-cap: "Number of active bear and moose hunters in the southeast region from 2006-2012"
+
+#create line graph 
+ggplot()+
+  geom_line(data = se_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
+  geom_line(data = se_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+  theme_minimal()
+
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-northhunters
+#| fig-cap: "Number of active bear and moose hunters in the northern region from 2006-2023"
+
+#create line graph 
+ggplot()+
+  geom_line(data = north_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
+  geom_line(data = north_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+  theme_minimal()
+
+```
+
+## Active Hunters vs. Total Harvest
+
+By examining the number of hunters against the number of harvests for both moose (@fig-harvesthuntermoose) and black bears (@fig-harvesthuntersbear) we can notice a few trends.
+Overall, for both moose and black bears in all three regions, as the number of hunters increases as does the total harvest.
+However, there are still many fluctuations in this trend, especially in the southeastern region.
+For both moose and black bear there are numerous dips and peaks where, as the number of hunters increase, the total harvest is significantly less or more than the overall average.
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-harvesthuntermoose
+#| fig-cap: "Number of active moose hunters plotted against total moose harvests from 2006-2023 in the northern (A), southeastern (B), and southwestern (C) region"
+
+plot1 <- 
+  ggplot()+
+  geom_line(data = north_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  theme_minimal()+
+  labs(tag = "A")
+
+plot2 <-
+  ggplot()+
+  geom_line(data = se_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  theme_minimal() +
+  labs(tag = "B")
+
+plot3 <-
+  ggplot()+
+  geom_line(data = sw_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 70))
+  labs(tag = "C")
+
+grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2)
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-harvesthuntersbear
+#| fig-cap: "Number of active bear hunters plotted against total bear harvests from 2012-2023 in the northern (A), southeastern (B), and southwestern (C) region"
+
+plot1 <- 
+  ggplot()+
+  geom_line(data = north_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  theme_minimal()+
+  labs(tag = "A")
+
+plot2 <-
+  ggplot()+
+  geom_line(data = se_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  theme_minimal() +
+  labs(tag = "B")
+
+plot3 <-
+  ggplot()+
+  geom_line(data = sw_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  theme_minimal() +
+  labs(tag = "C")
+
+grid.arrange(plot1, plot2, plot3, ncol = 2, nrow = 2)
+```
+
+We can also compare the number of hunters against the number of harvests for both moose (@fig-moose) and black bear (@fig-bear) for each WMU.
+From both graphs we can see that the more northern the WMUs tend to have higher harvest counts for lower hunter counts.
+Whereas, when the number of hunters is lower in the more southern regions, the less harvests there are.
+Overall, the more hunters the more harvests collected.
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-moose
+#| fig-cap: "Number of moose harvested in each WMU from 2006-2023"
+
+#create line graph 
+ggplot(all_moose, aes(x = `Active Hunters`, y = `Total Harvest`, color = WMU))+
+scale_color_distiller(palette = "RdPu")+
+geom_point(show.legend = TRUE)+
+theme_minimal()
+```
+
+```{r}
+#| echo: false
+#| warning: false
+#| label: fig-bear
+#| fig-cap: "Number of bear harvested in each WMU from 2012-2023"
+
+#create line graph 
+ggplot(all_bear, aes(x = `Active Hunters`, y = Harvest, color = WMU))+
+geom_point(show.legend = TRUE)+
+scale_color_distiller(palette = "YlGnBu")+
+theme_minimal()
+```
+
+# Discussion {#sec-discussion}
 

### 2024-04-20 scripts/02-data_cleaning.R

@@ -85,6 +85,19 @@ southwest_data_moose <-
             `Calf Harvest`= sum(`Calf Harvest`), 
             `Total Harvest`= sum(`Total Harvest`))
 
+#Creates a tibble of all the bull, calf, and cow harvests over the years from all regions
+types_data_moose <- 
+  raw_data_moose |>
+  group_by(Year)|>
+  summarise(`Bull Harvest`= sum(`Bull Harvest`), 
+            `Cow Harvest`= sum(`Cow Harvest`), 
+            `Calf Harvest`= sum(`Calf Harvest`))
+
+#creates a tibble of all moose WMU data, not grouped by year
+all_data_moose <- 
+  raw_data_moose |>
+  filter(WMU >= 1 & WMU <= 95)
+
 ## Bear tibbles ##
 #creates a tibble of the north region bear data by year
 north_data_bear <- 
@@ -110,6 +123,11 @@ southwest_data_bear <-
   summarise(`Active Hunters`= sum(`Active Hunters`), 
             Harvest= sum(Harvest))
 
+#creates a tibble of all bear WMU data, not grouped by year
+all_data_bear <- 
+  raw_data_bear |>
+  filter(WMU >= 1 & WMU <= 95)
+
 #### SAVE DATA ####
 write_csv(
   x = north_data_moose,
@@ -127,6 +145,16 @@ write_csv(
 )
 
 write_csv(
+  x = all_data_moose,
+  file = "data/analysis_data/all_data_moose.csv"
+)
+
+write_csv(
+  x = types_data_moose,
+  file = "data/analysis_data/types_data_moose.csv"
+)
+
+write_csv(
   x = north_data_bear,
   file = "data/analysis_data/north_data_bear.csv"
 )
@@ -141,3 +169,7 @@ write_csv(
   file = "data/analysis_data/southwest_data_bear.csv"
 )
 
+write_csv(
+  x = all_data_bear,
+  file = "data/analysis_data/all_data_bear.csv"
+)

### 2024-04-21 paper/paper.qmd
Finished the intro of the quarto document and started the discussion section

@@
 The analysis examines harvest counts from 2012 to 2023 across northern, southeastern, and southwestern Ontario.
-It was found that..."
+It was found that moose harvests have decreased over the past 20 years whereas black bear hunting has stayed relativly consistant.
+The implications and causiontions of these trends are important for species population in Ontario as well as the Ontario hunting industry."
 format: pdf
@@
 Alongside over hunting, having too large a population of a certain species can also put a strain on the ecosystem in which case hunting is an important method of regulating population.
-This paper explores the trends of active hunter and harvest numbers from 2006-2023 for moose and 2021-2023 for black bears.
-
-WHAT WAS FOUND
+The Ontario hunting industry is a significant economic contributor and, therefore, it is important to the province for hunting practices to continue.
+This paper examines these hunting activities in Ontario and how they have changed over the past few years.
+Specifically focusing on two majorly hunted species, moose and black bear, which are favoured by trophy hunters.
+Using data from the Ontario Data Catalogue active hunter and harvest numbers from 2006-2023 for moose and 2021-2023 for black bears are analyzed.
+From this analysis, it was found that moose harvests have drastically decreased since 2006 as has the number of active moose hunters.
+In comparison, black bear harvesting and hunting has stayed relatively consistent.
+While moose hunting has decreased there is still a significant population of moose hunters, especially compared to black bear hunters.
 
 The estimand of this paper is the correlation between the number of active hunters in Ontario and the number of annual moose and black bear harvests.
-Specifically, we examine how these numbers vary with regard to region and time
+Specifically, we examine how these numbers vary with regard to region and time.
 
@@
 Additionally, it identifies possible biases and factors that may have effected the analysis.
-
-MORE ABOUT THE SECTIONS
 
@@
 #| label: fig-harvestmoosetypes
-#| fig-cap: "Number of cow, bull, and calf moose types harvested between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+#| fig-cap: "Total number of cow, bull, and calf moose types harvested between 2012 and 2023"
 
@@
 
-TABLE?
-
 # Results {#sec-results}
 
-## Total Harvests
+## Total Harvest Trends
 
@@
 
-## Active Hunters
+By examining the data sets we also learn that for black bears the average harvest between 2012 and 2023 is 3,160 in the northern region, 2,046 in the southeastern region, and 177 in the southwestern region.
+In compression, for moose between 2006 and 2023 it is 3,356 in the northern region, 1,072 in the southeastern region, and 129 in the southwestern region.
+The maximum number of moose harvests was in 2006 with 7,342 total harvests.The minimum number of moose harvests was in 2023 with 2,915 total harvests.
+Whereas, for black bears the maximum number of harvests was in 2016 with 8,152 harvests.
+The minimum was in 2021 with 3,785 active hunters.
+
+## Active Hunter Trends
 
@@
 
+By examining the data sets we also learn that for black bears the average number of active hunters between 2012 and 2023 is 12,736 in the northern region, 12,717 in the southeastern region, and 499 in the southwestern region.
+In compression, for moose between 2006 and 2023 it is 49,666 in the northern region, 19,298 in the southeastern region, and 1332 in the southwestern region.
+The maximum number of active moose hunters was in 2010 with 88,369 total hunters.
+The minimum number of active hunters was in 2023 with 36,069 total hunters.
+Whereas, for black bears the maximum number of hunters was in 2016 with 31,480 hunters.
+The minimum was in 2013 with 20,891 active hunters.
+
 ## Active Hunters vs. Total Harvest
@@
 By examining the number of hunters against the number of harvests for both moose (@fig-harvesthuntermoose) and black bears (@fig-harvesthuntersbear) we can notice a few trends.
-Overall, for both moose and black bears in all three regions, as the number of hunters increases as does the total harvest.
+Overall, for both moose and black bears in all three regions, there is a linear relationship between the number of hunters and the number of harvests.
+From the grpahs we can see that as the number of hunters increases as does the total harvest.
 However, there are still many fluctuations in this trend, especially in the southeastern region.
@@
   theme_minimal()+
+  theme(axis.text.x = element_text(angle = 70))+
   labs(tag = "A")
@@
   theme_minimal() +
+  theme(axis.text.x = element_text(angle = 70))+
   labs(tag = "B")
@@
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 70))
+  theme(axis.text.x = element_text(angle = 70))+
   labs(tag = "C")
@@
   theme_minimal() +
+  theme(axis.text.x = element_text(angle = 70))+
   labs(tag = "C")
@@
 
+Overall, from all of these graphs we can see that every year less and less people are actively hunting moose.
+The number of people hunting moose and the number of moose being harvested is decreasing.
+Whereas, hunting black bears peaked greatly in 2016 but has since gone back to its pre 2016 harvest levels and appears to be staying relatively consistent.
+Additionally, we can see that while there are often more moose hunters than black bear hunters, more black bears are harvested yearly then moose on average.
+
 # Discussion {#sec-discussion}
 
-## What was done
+This paper has analyzed the correlation between the number of active hunters and the number of moose and black bear harvests in Ontario and the impact of time and region on these numbers.
+Additionally it examined the number of harvests against the number of active hunters.
+The analysis focused on three main regions, the northern, southeastern, and southwestern areas of Ontario.
+This was done with the analysis of two data sets obtained from the Ontario Data Catalogue.
 
-## Something we learned about the world
+## Ontario moose hunting is declining
 
-## Another thing we learned about the world
+## Black bears are popular hunting targets in Ontario
 
 ## Weaknesses
+
+The primary weakness of this analysis is its lack of depth.
+This is because it only focuses on and uses data with two main variables, number of active hunters and harvests.
+From the data alone it is difficult to know the exact reasons for certain trends, such as the decrease in moose harvesting without external research.
+This analysis could have been made stronger if paired with moose and black bear population data.
+More types of data and variables would have added more depth to this analysis.
+
+Another weakness is that the data doesn't cover a very large time span.
+The black bear data is only from 2012 to 2023 whereas the moose data covers a larger time frame from 2006 to 2023.
+Analyzing a data set with a larger range would provide more depth.
+This would also be beneficial as trends related to populations provide more insight when analyzed over a large period of time.
+The moose data only shows decreasing harvest rates so it would be interesting to see when they began increasing.
 

### 2024-04-22 paper/paper.qmd
Finished simulation and data testing scripts as well as running Lintr on all scripts

@@
 There are a total of 95 different WMUs which are grouped into three larger regions, northern, southeastern, and southwestern Ontario.
+However, the black bear data set only has data from WMUs 1 through 82 and the moose data set has data from WMUs 1-65.
 The second variable, *Year*, indicates the harvest year.
@@
 plot1 <- 
-  ggplot()+
-  geom_line(data = north_moose, aes(x = Year, y = `Active Hunters`), color = "black") +
-  theme_minimal()+
+  ggplot() +
+  geom_line(data = north_moose, aes(x = Year, y = `Active Hunters`),
+            color = "black") +
+  theme_minimal() +
   labs(tag = "A")
@@
 plot2 <-
-  ggplot()+
-  geom_line(data = se_moose, aes(x = Year, y = `Active Hunters`), color = "black") +
+  ggplot() +
+  geom_line(data = se_moose, aes(x = Year, y = `Active Hunters`),
+            color = "black") +
   theme_minimal() +
@@
 plot3 <-
-  ggplot()+
-  geom_line(data = sw_moose, aes(x = Year, y = `Active Hunters`), color = "black") +
+  ggplot() +
+  geom_line(data = sw_moose, aes(x = Year, y = `Active Hunters`),
+            color = "black") +
   theme_minimal() +
@@
 plot1 <- 
-  ggplot()+
-  geom_line(data = types_moose, aes(x = Year, y = `Cow Harvest`), color = "black") +
+  ggplot() +
+  geom_line(data = types_moose, aes(x = Year, y = `Cow Harvest`),
+            color = "black") +
   theme_minimal()+
@@
 plot2 <-
-  ggplot()+
-  geom_line(data = types_moose, aes(x = Year, y = `Calf Harvest`), color = "black") +
+  ggplot() +
+  geom_line(data = types_moose, aes(x = Year, y = `Calf Harvest`),
+            color = "black") +
   theme_minimal() +
@@
 plot3 <-
-  ggplot()+
-  geom_line(data = types_moose, aes(x = Year, y = `Bull Harvest`), color = "black") +
+  ggplot() +
+  geom_line(data = types_moose, aes(x = Year, y = `Bull Harvest`),
+            color = "black") +
   theme_minimal() +
@@
 plot1 <- 
-  ggplot()+
-  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "black") +
-  theme_minimal()+
+  ggplot() +
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`),
+            color = "black") +
+  theme_minimal() +
   labs(tag = "A")
@@
 plot2 <-
-  ggplot()+
-  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "black") +
+  ggplot() +
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`),
+            color = "black") +
   theme_minimal() +
@@
 plot3 <-
-  ggplot()+
-  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "black") +
+  ggplot() +
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`),
+            color = "black") +
   theme_minimal() +
@@
 plot1 <- 
-  ggplot()+
-  geom_line(data = north_bear, aes(x = Year, y = `Active Hunters`), color = "black") +
-  theme_minimal()+
+  ggplot() +
+  geom_line(data = north_bear, aes(x = Year, y = `Active Hunters`),
+            color = "black") +
+  theme_minimal() +
   labs(tag = "A")
@@
 plot2 <-
-  ggplot()+
-  geom_line(data = se_bear, aes(x = Year, y = `Active Hunters`), color = "black") +
+  ggplot() +
+  geom_line(data = se_bear, aes(x = Year, y = `Active Hunters`),
+            color = "black") +
   theme_minimal() +
@@
 plot3 <-
-  ggplot()+
-  geom_line(data = sw_bear, aes(x = Year, y = `Active Hunters`), color = "black") +
+  ggplot() +
+  geom_line(data = sw_bear, aes(x = Year, y = `Active Hunters`),
+            color = "black") +
   theme_minimal() +
@@
 plot1 <- 
-  ggplot()+
-  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "black") +
-  theme_minimal()+
+  ggplot() +
+  geom_line(data = north_bear, aes(x = Year, y = Harvest),
+            color = "black") +
+  theme_minimal() +
   labs(tag = "A")
@@
 plot2 <-
-  ggplot()+
-  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "black") +
+  ggplot() +
+  geom_line(data = se_bear, aes(x = Year, y = Harvest),
+            color = "black") +
   theme_minimal() +
@@
 plot3 <-
-  ggplot()+
-  geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "black") +
+  ggplot() +
+  geom_line(data = sw_bear, aes(x = Year, y = Harvest),
+            color = "black") +
   theme_minimal() +
@@
 #create line graph 
-ggplot()+
+ggplot() +
   geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "blue") + 
-  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`),
+            color = "red") +
   theme_minimal()
@@
 #create line graph 
-ggplot()+
+ggplot() +
   geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "blue") + 
-  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`),
+            color = "red") +
   theme_minimal()
@@
 #create line graph 
-ggplot()+
+ggplot() +
   geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "blue") + 
-  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`),
+            color = "red") +
   theme_minimal()
@@
 #create line graph 
-ggplot()+
+ggplot() +
   geom_line(data = sw_bear, aes(x = Year, y = Harvest), color = "blue") + 
-  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "red")+
-  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "green")+
+  geom_line(data = se_bear, aes(x = Year, y = Harvest), color = "red") +
+  geom_line(data = north_bear, aes(x = Year, y = Harvest), color = "green") +
   theme_minimal()
@@
 #create line graph 
-ggplot()+
-  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`), color = "blue") + 
-  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`), color = "red")+
-  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`), color = "green")+
+ggplot() +
+  geom_line(data = sw_moose, aes(x = Year, y = `Total Harvest`),
+            color = "blue") + 
+  geom_line(data = se_moose, aes(x = Year, y = `Total Harvest`),
+            color = "red") +
+  geom_line(data = north_moose, aes(x = Year, y = `Total Harvest`),
+            color = "green") +
   theme_minimal()
@@
 #create line graph 
-ggplot()+
-  geom_line(data = sw_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
-  geom_line(data = sw_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+ggplot() +
+  geom_line(data = sw_bear, aes(x = Year, y = `Active Hunters`),
+            color = "blue") +
+  geom_line(data = sw_moose, aes(x = Year, y = `Active Hunters`),
+            color = "red") +
   theme_minimal()
@@
 #create line graph 
-ggplot()+
-  geom_line(data = se_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
-  geom_line(data = se_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+ggplot() +
+  geom_line(data = se_bear, aes(x = Year, y = `Active Hunters`),
+            color = "blue") +
+  geom_line(data = se_moose, aes(x = Year, y = `Active Hunters`),
+            color = "red") +
   theme_minimal()
@@
 #create line graph 
-ggplot()+
-  geom_line(data = north_bear, aes(x = Year, y = `Active Hunters`), color = "blue") +
-  geom_line(data = north_moose, aes(x = Year, y = `Active Hunters`), color = "red")+
+ggplot() +
+  geom_line(data = north_bear, aes(x = Year, y = `Active Hunters`),
+            color = "blue") +
+  geom_line(data = north_moose, aes(x = Year, y = `Active Hunters`),
+            color = "red") +
   theme_minimal()
@@
 plot1 <- 
-  ggplot()+
-  geom_line(data = north_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
-  theme_minimal()+
-  theme(axis.text.x = element_text(angle = 70))+
+  ggplot() +
+  geom_line(data = north_moose, aes(x = `Active Hunters`, y = `Total Harvest`),
+            color = "red") +
+  theme_minimal() +
+  theme(axis.text.x = element_text(angle = 70)) +
   labs(tag = "A")
@@
 plot2 <-
-  ggplot()+
-  geom_line(data = se_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  ggplot() +
+  geom_line(data = se_moose, aes(x = `Active Hunters`, y = `Total Harvest`),
+            color = "red") +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 70))+
+  theme(axis.text.x = element_text(angle = 70)) +
   labs(tag = "B")
@@
 plot3 <-
-  ggplot()+
-  geom_line(data = sw_moose, aes(x = `Active Hunters`, y = `Total Harvest`), color = "red") +
+  ggplot() +
+  geom_line(data = sw_moose, aes(x = `Active Hunters`, y = `Total Harvest`),
+            color = "red") +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 70))+
+  theme(axis.text.x = element_text(angle = 70)) +
   labs(tag = "C")
@@
 plot1 <- 
-  ggplot()+
-  geom_line(data = north_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  ggplot() +
+  geom_line(data = north_bear, aes(x = `Active Hunters`, y = Harvest),
+            color = "blue") +
   theme_minimal()+
-  theme(axis.text.x = element_text(angle = 70))+
+  theme(axis.text.x = element_text(angle = 70)) +
   labs(tag = "A")
@@
 plot2 <-
-  ggplot()+
-  geom_line(data = se_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  ggplot() +
+  geom_line(data = se_bear, aes(x = `Active Hunters`, y = Harvest),
+            color = "blue") +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 70))+
+  theme(axis.text.x = element_text(angle = 70)) +
   labs(tag = "B")
@@
 plot3 <-
-  ggplot()+
-  geom_line(data = sw_bear, aes(x = `Active Hunters`, y = Harvest), color = "blue") +
+  ggplot() +
+  geom_line(data = sw_bear, aes(x = `Active Hunters`, y = Harvest),
+            color = "blue") +
   theme_minimal() +
-  theme(axis.text.x = element_text(angle = 70))+
+  theme(axis.text.x = element_text(angle = 70)) +
   labs(tag = "C")
@@
 #create line graph 
-ggplot(all_moose, aes(x = `Active Hunters`, y = `Total Harvest`, color = WMU))+
-scale_color_distiller(palette = "RdPu")+
-geom_point(show.legend = TRUE)+
+ggplot(all_moose, aes(x = `Active Hunters`, y = `Total Harvest`, color = WMU)) +
+scale_color_distiller(palette = "RdPu") +
+geom_point(show.legend = TRUE) +
 theme_minimal()
@@
 #create line graph 
-ggplot(all_bear, aes(x = `Active Hunters`, y = Harvest, color = WMU))+
-geom_point(show.legend = TRUE)+
-scale_color_distiller(palette = "YlGnBu")+
+ggplot(all_bear, aes(x = `Active Hunters`, y = Harvest, color = WMU)) +
+geom_point(show.legend = TRUE) +
+scale_color_distiller(palette = "YlGnBu") +
 theme_minimal()

### 2024-04-22 scripts/00-simulate_data.R

@@ -1,10 +1,10 @@
 #### Preamble ####
-# Purpose: Simulates the moose, black bear, white tailed deer, wolf and coyote harvest data from the Ontario Data Catalogue
+# Purpose: Simulates the moose and black bear data sets from
+# the Ontario Data Catalogue
 # Author: Samantha Barfoot
 # Date: 28 March 2024
 # Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
 
 #### Workspace setup ####
 library(tidyverse)
@@ -13,43 +13,304 @@ library(janitor)
 #### Simulate data ####
 set.seed(1)
 
-#variable that represents the number of columns in the simulated data set
-num_samples <- 1700
-
-
-# Black bear tailed deer simulated data #
-#simulated tibble table
-simulated_data_bear <-
+# Black bear simulated data
+#simulated tibble tables
+# Random values are generated for each column that is within a reasonable range
+north_data_bear <-
   tibble(
-    # Randomly pick an option, with replacement, 153 times
-    "WMU" = c(""),
     "Year" = sample(
-      x = 2009:2023,
-      size = num_samples,
+      x = 2012:2023,
+      size = 12,
       replace = TRUE
     ),
     "Active Hunters" = sample(
       x = 0:500,
-      size = num_samples,
+      size = 12,
       replace = TRUE
     ),
     "Harvest" = sample(
       x = 0:200,
-      size = num_samples,
+      size = 12,
+      replace = TRUE
+    )
+  )
+
+se_data_bear <-
+  tibble(
+    "Year" = sample(
+      x = 2012:2023,
+      size = 12,
+      replace = TRUE
+    ),
+    "Active Hunters" = sample(
+      x = 0:500,
+      size = 12,
+      replace = TRUE
+    ),
+    "Harvest" = sample(
+      x = 0:200,
+      size = 12,
+      replace = TRUE
+    )
+  )
+
+sw_data_bear <-
+  tibble(
+    "Year" = sample(
+      x = 2012:2023,
+      size = 12,
+      replace = TRUE
+    ),
+    "Active Hunters" = sample(
+      x = 0:500,
+      size = 12,
+      replace = TRUE
+    ),
+    "Harvest" = sample(
+      x = 0:200,
+      size = 12,
[…]

### 2024-04-22 scripts/02-data_cleaning.R

@@ -1,21 +1,22 @@
 #### Preamble ####
-# Purpose: Cleans the raw plane data recorded by two observers..... [...UPDATE THIS...]
+# Purpose: Cleans the raw data files obtained from the Ontario Data Catalogue
 # Author: Samantha Barfoot
 # Date: 28 March 2024
 # Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
-# Any other information needed?
+# Pre-requisites: download data from Ontario Data Catalogue
 
 #### Workspace setup ####
 library(tidyverse)
 library(MASS)
 library(stringr)
 library(dplyr)
+library(arrow)
+library(lintr)
 
 #### Read in data ####
 raw_data_moose <- read_csv("data/raw_data/moose_2023.csv",
-                          show_col_types = FALSE)
+                           show_col_types = FALSE)
 
 raw_data_bear <- read_csv("data/raw_data/black_bear_2023.csv",
                           show_col_types = FALSE)
@@ -25,106 +26,108 @@ num_records_moose <- dim(raw_data_moose)[1]
 num_records_bear <- dim(raw_data_bear)[1]
 
 #cleans WMU names so that there are no letters or beginning zeros for moose
-for(i in 1:num_records_moose){
-  unit <- raw_data_moose[[i,1]]
-  #Checks if the wmu value has a zero at the start or a letter at the end of the string and removes them
-  if(str_starts(unit,"0")){
-    unit <- str_remove(unit,"0")
+for (i in 1:num_records_moose){
+  unit <- raw_data_moose[[i, 1]]
+  #Checks if the wmu value has a zero at the start or a letter at the
+  # end of the string and removes them
+  if (str_starts(unit, "0")) {
+    unit <- str_remove(unit, "0")
   }
-  if(str_ends(unit,"[A-Za-z]")){
-    unit <- str_remove(unit,"[A-Za-z]")
+  if (str_ends(unit, "[A-Za-z]")) {
+    unit <- str_remove(unit, "[A-Za-z]")
   }
-  raw_data_moose[[i,1]] <- unit
+  raw_data_moose[[i, 1]] <- unit
 }
 
 #cleans WMU names so that there are no letters or beginning zeros  for bears
-for(i in 1:num_records_bear){
[…]

### 2024-04-22 scripts/03-test_data.R

@@ -1,15 +1,232 @@
 #### Preamble ####
-# Purpose: Tests... [...UPDATE THIS...]
+# Purpose: Tests data for values containing extra zeros and letters as well as
+# making sure the data is within a reasonable range (no bad data). Additionally,
+# it checks that "total" values are correct when adding up other values
 # Author: Samantha Barfoot
 # Date: 28 March 2024
 # Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: 
-# Any other information needed?
-
+# Pre-requisites: download data sets from the Ontario Data Catalogue
+# and run 02-data_cleaning.R
 
 #### Workspace setup ####
 library(tidyverse)
-# [...UPDATE THIS...]
+library(lintr)
+
+#### Read in data ####
+
+north_bear <-
+  read_csv(
+    file = "./data/analysis_data/north_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+se_bear <-
+  read_csv(
+    file = "./data/analysis_data/southeast_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+sw_bear <-
+  read_csv(
+    file = "./data/analysis_data/southwest_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+north_moose <-
+  read_csv(
+    file = "./data/analysis_data/north_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+se_moose <-
+  read_csv(
+    file = "./data/analysis_data/southeast_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+sw_moose <-
+  read_csv(
+    file = "./data/analysis_data/southwest_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+all_moose <-
+  read_csv(
+    file = "./data/analysis_data/all_data_moose.csv",
+    show_col_types = FALSE
+  )
+
+all_bear <-
+  read_csv(
+    file = "./data/analysis_data/all_data_bear.csv",
+    show_col_types = FALSE
+  )
+
+types_moose <-
+  read_csv(
+    file = "./data/analysis_data/types_data_moose.csv",
+    show_col_types = FALSE
+  )
 
 #### Test data ####
+
+#TEST 1: Check that the WMU values do not end in letters, only numbers
[…]

### 2024-04-23 paper/datasheet_template.qmd
Finished quarto paper write up, started data sheet, and ran lintr on scripts again

---
title: "Datasheet for 'Wildlife Hunting' data"
subtitle: "My subtitle if needed"
author: Samantha Barfoot
thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/wildlife_hunting.git"
date: today
date-format: long
abstract: "First sentence. Second sentence. Third sentence. Fourth sentence."
format: pdf
number-sections: true
bibliography: references.bib
---

Extract of the questions from @gebru2021datasheets.

**Motivation**

1. *For what purpose was the dataset created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The data set was created to enable analysis of Ontario moose and black bear hunting activity. Specifically, the number of active hunters and harvests of both moose and black bears by Wildlife Managment Region
2. *Who created the dataset (for example, which team, research group) and on behalf of which entity (for example, company, institution, organization)?*
    - The creator of the original data set are unknown but they were created on behalf of the Ministry of Resources and Forestry. These data sets were then modified by Samantha Barfoot to use in this paper.
3. *Who funded the creation of the dataset? If there is an associated grant, please provide the name of the grantor and the grant name and number.*
    - The creation of the original data set was funded by the goverment of Ontario

**Composition**

1. *What do the instances that comprise the dataset represent (for example, documents, photos, people, countries)? Are there multiple types of instances (for example, movies, users, and ratings; people and interactions between them; nodes and edges)? Please provide a description.*
	- TBD
2. *How many instances are there in total (of each type, if appropriate)?*
	- TBD
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- TBD
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- TBD
5. *Is there a label or target associated with each instance? If so, please provide a description.*
	- TBD
6. *Is any information missing from individual instances? If so, please provide a description, explaining why this information is missing (for example, because it was unavailable). This does not include intentionally removed information, but might include, for example, redacted text.*
	- TBD
7. *Are relationships between individual instances made explicit (for example, users' movie ratings, social network links)? If so, please describe how these relationships are made explicit.*
	- TBD
8. *Are there recommended data splits (for example, training, development/validation, testing)? If so, please provide a description of these splits, explaining the rationale behind them.*
	- TBD
9. *Are there any errors, sources of noise, or redundancies in the dataset? If so, please provide a description.*
	- TBD
10. *Is the dataset self-contained, or does it link to or otherwise rely on external resources (for example, websites, tweets, other datasets)? If it links to or relies on external resources, a) are there guarantees that they will exist, and remain constant, over time; b) are there official archival versions of the complete dataset (that is, including the external resources as they existed at the time the dataset was created); c) are there any restrictions (for example, licenses, fees) associated with any of the external resources that might apply to a dataset consumer? Please provide descriptions of all external resources and any restrictions associated with them, as well as links or other access points, as appropriate.*
	- TBD
11. *Does the dataset contain data that might be considered confidential (for example, data that is protected by legal privilege or by doctor-patient confidentiality, data that includes the content of individuals' non-public communications)? If so, please provide a description.*
	- TBD
12. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety? If so, please describe why.*
	- TBD
13. *Does the dataset identify any sub-populations (for example, by age, gender)? If so, please describe how these subpopulations are identified and provide a description of their respective distributions within the dataset.*
	- TBD
14. *Is it possible to identify individuals (that is, one or more natural persons), either directly or indirectly (that is, in combination with other data) from the dataset? If so, please describe how.*
	- TBD
15. *Does the dataset contain data that might be considered sensitive in any way (for example, data that reveals race or ethnic origins, sexual orientations, religious beliefs, political opinions or union memberships, or locations; financial or health data; biometric or genetic data; forms of government identification, such as social security numbers; criminal history)? If so, please provide a description.*
	- TBD
16. *Any other comments?*
	- TBD

**Collection process**

1. *How was the data associated with each instance acquired? Was the data directly observable (for example, raw text, movie ratings), reported by subjects (for example, survey responses), or indirectly inferred/derived from other data (for example, part-of-speech tags, model-based guesses for age or language)? If the data was reported by subjects or indirectly inferred/derived from other data, was the data validated/verified? If so, please describe how.*
	- TBD
2. *What mechanisms or procedures were used to collect the data (for example, hardware apparatuses or sensors, manual human curation, software programs, software APIs)? How were these mechanisms or procedures validated?*
	- TBD
3. *If the dataset is a sample from a larger set, what was the sampling strategy (for example, deterministic, probabilistic with specific sampling probabilities)?*
	- TBD
4. *Who was involved in the data collection process (for example, students, crowdworkers, contractors) and how were they compensated (for example, how much were crowdworkers paid)?*
	- TBD
5. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances (for example, recent crawl of old news articles)? If not, please describe the timeframe in which the data associated with the instances was created.*
	- TBD
6. *Were any ethical review processes conducted (for example, by an institutional review board)? If so, please provide a description of these review processes, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
7. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources (for example, websites)?*
	- TBD
8. *Were the individuals in question notified about the data collection? If so, please describe (or show with screenshots or other information) how notice was provided, and provide a link or other access point to, or otherwise reproduce, the exact language of the notification itself.*
	- TBD
9. *Did the individuals in question consent to the collection and use of their data? If so, please describe (or show with screenshots or other information) how consent was requested and provided, and provide a link or other access point to, or otherwise reproduce, the exact language to which the individuals consented.*
	- TBD
10. *If consent was obtained, were the consenting individuals provided with a mechanism to revoke their consent in the future or for certain uses? If so, please provide a description, as well as a link or other access point to the mechanism (if appropriate).*
	- TBD
11. *Has an analysis of the potential impact of the dataset and its use on data subjects (for example, a data protection impact analysis) been conducted? If so, please provide a description of this analysis, including the outcomes, as well as a link or other access point to any supporting documentation.*
	- TBD
12. *Any other comments?*
	- TBD

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done (for example, discretization or bucketing, tokenization, part-of-speech tagging, SIFT feature extraction, removal of instances, processing of missing values)? If so, please provide a description. If not, you may skip the remaining questions in this section.*
	- TBD
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data (for example, to support unanticipated future uses)? If so, please provide a link or other access point to the "raw" data.*
	- TBD
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- TBD
4. *Any other comments?*
	- TBD

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- TBD
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- TBD
3. *What (other) tasks could the dataset be used for?*
	- TBD
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? For example, is there anything that a dataset consumer might need to know to avoid uses that could result in unfair treatment of individuals or groups (for example, stereotyping, quality of service issues) or other risks or harms (for example, legal risks, financial harms)? If so, please provide a description. Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- TBD
5. *Are there tasks for which the dataset should not be used? If so, please provide a description.*
	- TBD
6. *Any other comments?*
	- TBD

**Distribution**

1. *Will the dataset be distributed to third parties outside of the entity (for example, company, institution, organization) on behalf of which the dataset was created? If so, please provide a description.*
	- TBD
2. *How will the dataset be distributed (for example, tarball on website, API, GitHub)? Does the dataset have a digital object identifier (DOI)?*
	- TBD
3. *When will the dataset be distributed?*
	- TBD
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? If so, please describe this license and/ or ToU, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- TBD
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any relevant licensing terms, as well as any fees associated with these restrictions.*
	- TBD
6. *Do any export controls or other regulatory restrictions apply to the dataset or to individual instances? If so, please describe these restrictions, and provide a link or other access point to, or otherwise reproduce, any supporting documentation.*
	- TBD
7. *Any other comments?*
	- TBD

**Maintenance**

1. *Who will be supporting/hosting/maintaining the dataset?*
	- TBD
2. *How can the owner/curator/manager of the dataset be contacted (for example, email address)?*
	- TBD
3. *Is there an erratum? If so, please provide a link or other access point.*
	- TBD
4. *Will the dataset be updated (for example, to correct labeling errors, add new instances, delete instances)? If so, please describe how often, by whom, and how updates will be communicated to dataset consumers (for example, mailing list, GitHub)?*
	- TBD
5. *If the dataset relates to people, are there applicable limits on the retention of the data associated with the instances (for example, were the individuals in question told that their data would be retained for a fixed period of time and then deleted)? If so, please describe these limits and explain how they will be enforced.*
	- TBD
6. *Will older versions of the dataset continue to be supported/hosted/maintained? If so, please describe how. If not, please describe how its obsolescence will be communicated to dataset consumers.*
	- TBD
7. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so? If so, please provide a description. Will these contributions be validated/verified? If so, please describe how. If not, why not? Is there a process for communicating/distributing these contributions to dataset consumers? If so, please provide a description.*
	- TBD
8. *Any other comments?*
	- TBD

\newpage

# References

### 2024-04-23 paper/paper.qmd
Finished quarto paper write up, started data sheet, and ran lintr on scripts again

@@
 abstract: "This paper analyzes Ontario wildlife harvest counts, specifically moose and black bears, and the number of active hunters based on region.
-The analysis examines harvest counts from 2012 to 2023 across northern, southeastern, and southwestern Ontario.
-It was found that moose harvests have decreased over the past 20 years whereas black bear hunting has stayed relativly consistant.
-The implications and causiontions of these trends are important for species population in Ontario as well as the Ontario hunting industry."
+The analysis examines harvest counts from 2006 to 2023 across Northern, Southeastern, and Southwestern Ontario.
+It was found that moose harvests have decreased over the past 20 years whereas black bear hunting has stayed relatively consistent.
+The implications and causes of these trends are important for species population in Ontario as well as the Ontario hunting industry."
 format: pdf
@@
 library(gridExtra)
+library(arrow)
 ```
@@
 Every year the Ontario Ministry of Natural Resources and Forestry records data on the number of black bears and moose hunted per year and the number of active hunters per year.
-Ontario moose hunting season typically runs from early September to late October [@moose] and black bear season runs from early May to mid June or mid August to late November depending on the region [@bear].
-To hunt either of these animals, hunters must have the proper licence and are responsible for identifying their target and obtaining tags.
-Moose tags indicate what type of moose a hunter is allowed to harvest whether it be a cow, calf, or bull [@moose].
-Hunting in Ontario is a significant contributor to the Ontario economy and tourism industry.
+Ontario moose hunting season typically runs from early September to late October [@moose] and the black bear season runs from early May to mid-June and mid-August to late November depending on the region [@bear].
+To hunt either of these animals, hunters must have the proper license and are responsible for identifying their target and obtaining tags, which indicate how many harvests they can take.
+Moose tags also indicate what type of moose a hunter is allowed to harvest whether it be a cow, calf, or bull [@moose].
+Hunting is a significant contributor to the Ontario economy and tourism industry.
 It supports the employment of thousands of people and some years generates more money than the film and television industry in Ontario [@huntAs].
-One year Canadian hunters spent around \$1.2 billion on hunting trips with \$70 on the hunting licences alone [@huntAs].
+One year Canadian hunters spent around \$1.2 billion on hunting trips with \$70 million on the hunting licenses alone [@huntAs].
 In 2000, the hunting industry represented \$1.5 billion of the economy [@huntAs].
-Despite being such a large industry there is still a significant amount of poaching and other illegal hunting practices performed in Ontario [@poachNum].
-In 2016 there were 384 major offences related to these practices [@poachNum].
+Despite being such a large industry, there is still a significant amount of poaching and other illegal hunting practices performed in Ontario [@poachNum].
+In 2016, there were 384 major offences related to these practices with people paying thousands of dollars in fines [@poachNum].
 
-Therefore, it is important to properly regulate and monitor the wildlife populations as well as the hunting and harvest numbers so that species are not put at risk by over hunting.
+Therefore, it is important to properly regulate and monitor the wildlife populations as well as the hunting and harvest numbers so that species are not put at risk by overhunting.
 Additionally, it is important to analyze how these levels change over the years and what might cause these changes.
-Alongside over hunting, having too large a population of a certain species can also put a strain on the ecosystem in which case hunting is an important method of regulating population.
+Alongside overhunting, having too large a population of a certain species can also put a strain on the ecosystem, in which case hunting can be an important method of regulating population.
 The Ontario hunting industry is a significant economic contributor and, therefore, it is important to the province for hunting practices to continue.
@@
 Specifically focusing on two majorly hunted species, moose and black bear, which are favoured by trophy hunters.
-Using data from the Ontario Data Catalogue active hunter and harvest numbers from 2006-2023 for moose and 2021-2023 for black bears are analyzed.
+Using data from the Ontario Data Catalogue active hunter and harvest numbers from 2006-2023 for moose and 2012-2023 for black bears are analyzed.
 From this analysis, it was found that moose harvests have drastically decreased since 2006 as has the number of active moose hunters.
-In comparison, black bear harvesting and hunting has stayed relatively consistent.
-While moose hunting has decreased there is still a significant population of moose hunters, especially compared to black bear hunters.
+In comparison, black bear harvesting and hunting have stayed relatively consistent.
+While moose hunting has decreased, there is still a significant population of moose hunters, especially compared to black bear hunters.
 
@@
 
-@sec-data of this paper covers the broader context of the Ontario moose and black bear hunting activty and harvest data sets from the Ontario Data Catalogue.
-It examines the variables used in the data set using graphs to help explain them as well as discussing high-level aspects of the data cleaning process.
-@sec-results further examines the relationship between the variables and the finding from the analysis of the two data sets.
-The @sec-discussion section of this paper explores what was learned from this analysis and the implications it has on Ontario's hunting industry and environment.
-Additionally, it identifies possible biases and factors that may have effected the analysis.
+@sec-data of this paper covers the broader context of the Ontario moose and black bear hunting activity and harvest data sets from the Ontario Data Catalogue.
+It examines the variables used in the data set using graphs to help explain them as well as discusses high-level aspects of the data cleaning process.
+@sec-results further examines the relationship between the variables and the findings from the analysis of the two data sets.
+@sec-discussion of this paper explores what was learned from this analysis and the implications it has on Ontario's hunting industry and environment as well as weaknesses in the analysis.
 
-The graphs and tables in this paper were made with [@citeR] using R studio.
-The creation of these graphs and tables were made with `ggplot` [@ggplotRef], `tiddyverse` [@tidyverseRef], `gridExtra` [@gridRef] packages.
-The analysis and cleaning of the data was conducted with and `dyplr` [@dplRef ], `MASS` [@massRef], `janitor` [@janitorRef] and `stringr` [@stringRef] packages.
+The graphs and tables in this paper were made with using R studio [@citeR].
+The creation of these graphs and tables were made with `ggplot` [@ggplotRef], `tidyverse` [@tidyverseRef], `gridExtra` [@gridRef] packages.
+The analysis and cleaning of the data was conducted with `dyplr` [@dplRef ], `MASS` [@massRef], `janitor` [@janitorRef], and `stringr` [@stringRef] packages.
 
@@
 north_bear <-
-  read_csv(
-    file = "../data/analysis_data/north_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/north_data_bear.parquet"
   )
@@
 se_bear <-
-  read_csv(
-    file = "../data/analysis_data/southeast_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/southeast_data_bear.parquet"
   )
@@
 sw_bear <-
-  read_csv(
-    file = "../data/analysis_data/southwest_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/southwest_data_bear.parquet"
   )
@@
 north_moose <-
-  read_csv(
-    file = "../data/analysis_data/north_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/north_data_moose.parquet"
   )
@@
 se_moose <-
-  read_csv(
-    file = "../data/analysis_data/southeast_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/southeast_data_moose.parquet"
   )
@@
 sw_moose <-
-  read_csv(
-    file = "../data/analysis_data/southwest_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/southwest_data_moose.parquet"
   )
@@
 all_moose <-
-  read_csv(
-    file = "../data/analysis_data/all_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/all_data_moose.parquet"
   )
@@
 types_moose <-
-  read_csv(
-    file = "../data/analysis_data/types_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/types_data_moose.parquet"
   )
@@
 all_bear <-
-  read_csv(
-    file = "../data/analysis_data/all_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "../data/analysis_data/all_data_bear.parquet"
   )
@@
 The data sets used in this report, *Moose hunting activity and harvests* and *Black bear hunting activity and harvests*, were obtained from the Ontario Data Catalogue.
-Specifically, the data is provided by the ministry of Natural Resources and Forestry and were last updated on February 16, 2024.
-They data sets provide the estimated hunter and harvest numbers by region across Ontario from 2006 for the moose data set and 2012 for the black bear data set.
+Specifically, the data is provided by the Ministry of Natural Resources and Forestry and was last updated on February 16, 2024.
+The data sets provide the estimated hunter and harvest numbers by region across Ontario from 2006 for the moose data set and 2012 for the black bear data set.
 The numbers in these data sets are based on replies that the Ontario government received from a sample of hunters across Ontario.
@@
 The number of active hunters and harvests could have been measured by surveys or the government asking all hunters registered for moose and black bear hunting how many harvests they collected.
-To legally hunt moose and black bears, hunters must register and obtain licences.
-Therefore, to measure how many active hunters are in Ontario the Ministry of Resources and Forestry might have counted how many registered hunters they have for each species.
-However, they can only estimate how many harvests were collected each year as it depends on hunters being honest about how many harvests they made and that all hunters posses the proper permits.
+To legally hunt moose and black bears, hunters must register and obtain licenses.
+Therefore, to measure how many active hunters are in Ontario, the Ministry of Resources and Forestry might have counted how many registered hunters they have for each species.
+The data collected are only estimations as it is composed of self-reported harvests.
+Hunters voluntarily submitted this data and thus might not entirely reflect the actual number of harvests.
+They can only estimate how many harvests were collected each year as it depends on hunters being honest about how many harvests they made and that all hunters possess the proper permits.
 
@@
 The first variable,*WMU*, is the Wildlife Management Unit which indicates the specific region in Ontario.
-There are a total of 95 different WMUs which are grouped into three larger regions, northern, southeastern, and southwestern Ontario.
-However, the black bear data set only has data from WMUs 1 through 82 and the moose data set has data from WMUs 1-65.
+There are a total of 95 different WMUs, which are grouped into three larger regions, Northern (see @fig-map1), Southeastern (see @fig-map2), and Southwestern (see @fig-map3) Ontario.
+However, the black bear data set only has data from WMUs 1 through 82 and the moose data set has data from WMUs 1 through 65.
 The second variable, *Year*, indicates the harvest year.
@@
 *Total Harvest* (see @fig-harvestmoose) is the sum of all bull, cow, and calf harvests for that specific WMU and year.
-The original data set for black bear hunting and harvests contains the same variables of WMU, year, and active hunters (see @fig-huntersbear).
-However, instead of bull, cow, calf, and total harvest there is only one harvest variable called *Harvest* (see @fig-harvestbear) as all black bears are grouped as one.
+The original data set for black bear hunting and harvests contain the same variables of WMU, year, and active hunters (see @fig-huntersbear).
+However, instead of bull, cow, calf, and total harvest, there is only one harvest variable called *Harvest* (see @fig-harvestbear) as all black bears are grouped as one.
+
+![Map of Northern Ontario region and the included WMUs (image credit: ontario.ca)](northern.png){#fig-map1}
+
+![Map of southeast Ontario region and the included WMUs (image credit: ontario.ca](southeast.png){#fig-map2}
+
+![Map of southwest Ontario region and the included WMUs (image credit: ontario.ca)](southwest.png){#fig-map3}
 
@@
 #| label: fig-huntersmoose
-#| fig-cap: "Number of active moose hunters between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+#| fig-cap: "Number of active moose hunters between 2012 and 2023 in the Northern (A), Southeastern (B), and Southwestern (C) region"
+#| fig.pos: "H"
 
@@
 #| fig-cap: "Total number of cow, bull, and calf moose types harvested between 2012 and 2023"
+#| fig.pos: "H"
 
@@
 #| label: fig-harvestmoose
-#| fig-cap: "Number of moose harvested between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+#| fig-cap: "Number of moose harvested between 2012 and 2023 in the Northern (A), Southeastern (B), and Southwestern (C) region"
+#| fig.pos: "H"
 
@@
 #| label: fig-huntersbear
-#| fig-cap: "Number of active black bear hunters between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+#| fig-cap: "Number of active black bear hunters between 2012 and 2023 in the Northern (A), Southeastern (B), and Southwestern (C) region"
+#| fig.pos: "H"
 
@@
 #| label: fig-harvestbear
-#| fig-cap: "Number of bears harvested between 2012 and 2023 in the northern (A), southeastern (B), and southwestern (C) region"
+#| fig-cap: "Number of bears harvested between 2012 and 2023 in the Northern (A), Southeastern (B), and Southwestern (C) region"
+#| fig.pos: "H"
 
@@
 
-To clean the data, each animals data sets were split into three sets, one for WMUs in the northern, southeastern, and southwestern region.
+To clean the data, each animal's data sets were split into three sets, one for WMUs in the Northern, Southeastern, and Southwestern regions.
 This way, different regions and their harvest and active hunter numbers could be plotted against each other.
@@
 Each cleaned data set also contains the same variables as the original data sets.
-Some cleaning was also applied to the WMU values as many of them had extra zeros and letter joined to them.
+Some cleaning was also applied to the WMU values as many of them had extra zeros and letters joined to them.
 
@@
 
-By analyzing each hunting region, northern, southeastern, and southwestern, separately, we can examine how the total harvest rates of black bears and moose compare.
-@fig-north shows that the harvest rates for moose generally decreases before 2018 whereas the amount of black bear harvests increases from 2012 to 2016 before decreasing and increasing again in 2021.
+By analyzing each hunting region separately, we can examine how the total harvest rates of black bears and moose compare.
+@fig-north shows that the harvest rates for moose generally decrease before 2018 whereas the amount of black bear harvests increases from 2012 to 2016 before decreasing and increasing again in 2021.
 The trend lines on @fig-north cross three times between 2012 and 2023 with black bear harvest rates sometimes being higher than moose rates and vice versa.
@@
 Whereas, in @fig-southwest the black bear harvest peak stays relatively consistent between 2016 and 2019 before greatly decreasing after 2016.
-In @fig-southwest, similar to @fig-southeast the number of black bear harvests is always higher than the number of moose harvests except in 2020 where they are equal.
-Interestingly, in all three graphs, the number of moose harvests tend to decrease overall between 2006 and 2023, most dramatically in the northern and southwestern regions.
-While all three of these graphs show that black bear harvest numbers are generally higher than moose, the number of harvests for each species are drastically different.
-In the northern region moose harvests reach around 5500 and black bears around 5000.
-Whereas, in the southeast region they reach 1500 and 3000, respectively, and in the southwestern region, 350 and 275 respectively.
+In @fig-southwest, similar to @fig-southeast, the number of black bear harvests is always higher than the number of moose harvests except in 2020 where they are equal.
+Interestingly, in all three graphs, the number of moose harvests tends to decrease overall between 2006 and 2023, most dramatically in the Northern and Southwestern regions.
+While all three of these graphs show that black bear harvest numbers are generally higher than moose, the number of harvests for each species is drastically different.
+In the Northern region, moose harvests reach around 5,500, and black bears around 5,000.
+Whereas, in the southeast region they reach 1,500 and 3,000, respectively, and in the Southwestern region, 350 and 275, respectively.
 
@@
 #| fig-cap: "Northern harvest numbers for moose and bears between 2006 and 2023"
+#| fig.pos: "H"
 
@@
 #| fig-cap: "Southeastern harvest numbers for moose and bears between 2006 and 2023"
+#| fig.pos: "H"
 
@@
 #| fig-cap: "Southwestern harvest numbers for moose and bears between 2006 and 2023"
+#| fig.pos: "H"
 
@@
 
+\newpage
 We can also compare the harvest rates of each species separately for each region.
-@fig-bearharvest we can see more clearly what we observed previously where the amount of black bears harvested is very different between each region.
-In the northern region, on average, has the most the most harvests followed by the southeast and then southwest region which is significantly lower than the other two regions.
-However, the northern and southeastern rates cross over at two points around 2020.
-@fig-mooseharvest shows a similar trend with the northern region always having the most harvests followed by the southeastern then the southwestern regions with no intersections.
-Both species likely experience this difference in harvest numbers because of the regions size.
-The northern region is significantly larger than the southeast and southwest regions.
-However, it is interesting that the northern and southeastern harvest rates are so similar between 2018 and 2023 given that the areas are of drastically different size.
+In @fig-bearharvest we can see more clearly what we observed previously where the amount of black bears harvested is very different between each region.
+The Northern region, on average, has the most harvests followed by the southeast and then the southwest region which is significantly lower than the other two regions.
+However, the Northern and Southeastern rates cross over at two points around 2020.
+@fig-mooseharvest shows a similar trend with the Northern region always having the most harvests followed by the Southeastern and then the Southwestern regions with no intersections.
+Both species likely experience this difference in harvest numbers because of the region's size as the Northern region is significantly larger than the southeast and southwest regions.
+However, it is interesting that the Northern and Southeastern harvest rates are so similar between 2018 and 2023 given that the areas are of drastically different sizes.
 
@@
 #| label: fig-bearharvest
-#| fig-cap: "Bear harvest numbers for the northern, southeastern and southwestern regions from 2012-2023"
+#| fig-cap: "Bear harvest numbers for the Northern, Southeastern and Southwestern regions from 2012-2023"
+#| fig.pos: "H"
 
@@
 #| label: fig-mooseharvest
-#| fig-cap: "Moose harvest numbers for the northern, southeastern and southwestern regions from 2006-2023"
+#| fig-cap: "Moose harvest numbers for the Northern, Southeastern and Southwestern regions from 2006-2023"
+#| fig.pos: "H"
 
@@
 
-By examining the data sets we also learn that for black bears the average harvest between 2012 and 2023 is 3,160 in the northern region, 2,046 in the southeastern region, and 177 in the southwestern region.
-In compression, for moose between 2006 and 2023 it is 3,356 in the northern region, 1,072 in the southeastern region, and 129 in the southwestern region.
-The maximum number of moose harvests was in 2006 with 7,342 total harvests.The minimum number of moose harvests was in 2023 with 2,915 total harvests.
-Whereas, for black bears the maximum number of harvests was in 2016 with 8,152 harvests.
-The minimum was in 2021 with 3,785 active hunters.
+By examining the data sets we also learn that for black bears the average harvest between 2012 and 2023 is 3,160 in the Northern region, 2,046 in the Southeastern region, and 177 in the Southwestern region.
+In comparison, for moose between 2006 and 2023, it is 3,356 in the Northern region, 1,072 in the Southeastern region, and 129 in the Southwestern region.
+The maximum number of moose harvests was in 2006 with 7,342 total harvests.
+The minimum number of moose harvests was in 2023 with 2,915 total harvests.
+Whereas, for black bears, the maximum number of harvests was in 2016 with 8,152 harvests.
+The minimum was in 2021 with 3,785 harvests.
 
@@
 However, the number of moose hunters has significantly decreased since 2006.
-In 2006 there was over 65,000 moose hunters and in 2023 there were only around 25,000 moose hunters.
+In 2006 there were over 65,000 moose hunters and in 2023 there were only around 25,000 moose hunters.
 In comparison, the number of black bear hunters has remained relatively consistent compared to moose hunters since 2012.
 The southwest region (see @fig-swhunters) has a similar trend with the number of moose hunters greatly decreasing since 2006.
-Additionally, there has generally been less bear hunters in the region compared to moose hunters up until 2021 at which point the number of black bear hunters surpassed the number of moose hunters.
-The southeast region (see @fig-sehunters), however, has a different trend; the number of moose hunters in the region stayed relatively consistent from 2006 up to 2019 after which it dramatically decreased.
+Additionally, there have generally been fewer bear hunters in the region compared to moose hunters up until 2021 at which point the number of black bear hunters surpassed the number of moose hunters.
+The southeast region (see @fig-sehunters), however, has a different trend; the number of moose hunters in the region stayed relatively consistent from 2006 up to 2019 after which it greatly decreased.
 Around this time the number of black bear hunters also surpassed the number of moose hunters.
 Since 2012 the number of black bear hunters has increased.
-The discrepancy between the number of hunters by regain could, again, be due to the fact that the southwestern and southeastern regions are much smaller than the northern region.
+The discrepancy between the number of hunters by regain could, again, be because the Southwestern and Southeastern regions are much smaller than the Northern region.
 
@@
 #| fig-cap: "Number of active bear and moose hunters in the southwest region from 2006-2023"
+#| fig.pos: "H"
 
@@
 #| fig-cap: "Number of active bear and moose hunters in the southeast region from 2006-2012"
+#| fig.pos: "H"
 
@@
 #| label: fig-northhunters
-#| fig-cap: "Number of active bear and moose hunters in the northern region from 2006-2023"
+#| fig-cap: "Number of active bear and moose hunters in the Northern region from 2006-2023"
+#| fig.pos: "H"
 
@@
 
-By examining the data sets we also learn that for black bears the average number of active hunters between 2012 and 2023 is 12,736 in the northern region, 12,717 in the southeastern region, and 499 in the southwestern region.
-In compression, for moose between 2006 and 2023 it is 49,666 in the northern region, 19,298 in the southeastern region, and 1332 in the southwestern region.
+\newpage
+By examining the data sets, we also learn that for black bears the average number of active hunters between 2012 and 2023 is 12,736 in the Northern region, 12,717 in the Southeastern region, and 499 in the Southwestern region.
+In comparison, for moose between 2006 and 2023, it is 49,666 in the Northern region, 19,298 in the Southeastern region, and 1332 in the Southwestern region.
 The maximum number of active moose hunters was in 2010 with 88,369 total hunters.
 The minimum number of active hunters was in 2023 with 36,069 total hunters.
-Whereas, for black bears the maximum number of hunters was in 2016 with 31,480 hunters.
+Whereas, for black bears, the maximum number of hunters was in 2016 with 31,480 hunters.
 The minimum was in 2013 with 20,891 active hunters.
@@
 
-By examining the number of hunters against the number of harvests for both moose (@fig-harvesthuntermoose) and black bears (@fig-harvesthuntersbear) we can notice a few trends.
-Overall, for both moose and black bears in all three regions, there is a linear relationship between the number of hunters and the number of harvests.
-From the grpahs we can see that as the number of hunters increases as does the total harvest.
-However, there are still many fluctuations in this trend, especially in the southeastern region.
-For both moose and black bear there are numerous dips and peaks where, as the number of hunters increase, the total harvest is significantly less or more than the overall average.
+By examining the number of harvests against the number of hunters for both moose (@fig-harvesthuntermoose) and black bears (@fig-harvesthuntersbear) we notice a few trends.
+Overall, for both moose and black bears in all three regions, there is an almost linear relationship between the number of hunters and the number of harvests.
+From the graphs, we can see that as the number of hunters increases as does the total harvest.
+However, there are still many fluctuations in this trend, especially in the Southeastern region.
+For both moose and black bears there are numerous dips and peaks where, as the number of hunters increases, the total harvest is significantly less or more than the overall average.
 
@@
 #| label: fig-harvesthuntermoose
-#| fig-cap: "Number of active moose hunters plotted against total moose harvests from 2006-2023 in the northern (A), southeastern (B), and southwestern (C) region"
+#| fig-cap: "Number of active moose hunters plotted against total moose harvests from 2006-2023 in the Northern (A), Southeastern (B), and Southwestern (C) region"
+#| fig.pos: "H"
 
@@
 #| label: fig-harvesthuntersbear
-#| fig-cap: "Number of active bear hunters plotted against total bear harvests from 2012-2023 in the northern (A), southeastern (B), and southwestern (C) region"
+#| fig-cap: "Number of active bear hunters plotted against total bear harvests from 2012-2023 in the Northern (A), Southeastern (B), and Southwestern (C) region"
+#| fig.pos: "H"
 
@@
 
-We can also compare the number of hunters against the number of harvests for both moose (@fig-moose) and black bear (@fig-bear) for each WMU.
-From both graphs we can see that the more northern the WMUs tend to have higher harvest counts for lower hunter counts.
-Whereas, when the number of hunters is lower in the more southern regions, the less harvests there are.
+We can also compare the number of harvests against the number of hunters for both moose (@fig-moose) and black bear (@fig-bear) for each WMU.
+From both graphs, we can see that the more Northern WMUs tend to have higher harvest counts for lower hunter counts.
+Whereas, when the number of hunters is lower in the more southern regions, the fewer harvests there are.
 Overall, the more hunters the more harvests collected.
+
+Overall, from all of these graphs, we can see that every year fewer and fewer people are actively hunting moose.
+The number of people hunting moose and the number of moose being harvested is decreasing.
+Whereas, hunting black bears peaked greatly in 2016 but has since gone back to its pre-2016 harvest levels and appears to be staying relatively consistent.
+Additionally, we can see that while there are often more moose hunters than black bear hunters, more black bears are harvested yearly than moose on average.
 
@@
 #| fig-cap: "Number of moose harvested in each WMU from 2006-2023"
+#| fig.pos: "H"
 
@@
 #| fig-cap: "Number of bear harvested in each WMU from 2012-2023"
+#| fig.pos: "H"
 
@@
 
-Overall, from all of these graphs we can see that every year less and less people are actively hunting moose.
-The number of people hunting moose and the number of moose being harvested is decreasing.
-Whereas, hunting black bears peaked greatly in 2016 but has since gone back to its pre 2016 harvest levels and appears to be staying relatively consistent.
-Additionally, we can see that while there are often more moose hunters than black bear hunters, more black bears are harvested yearly then moose on average.
-
 # Discussion {#sec-discussion}
@@
 This paper has analyzed the correlation between the number of active hunters and the number of moose and black bear harvests in Ontario and the impact of time and region on these numbers.
-Additionally it examined the number of harvests against the number of active hunters.
-The analysis focused on three main regions, the northern, southeastern, and southwestern areas of Ontario.
+Additionally, it examined the number of harvests against the number of active hunters.
+The analysis focused on three main regions, the Northern, Southeastern, and Southwestern areas of Ontario.
 This was done with the analysis of two data sets obtained from the Ontario Data Catalogue.
 
-## Ontario moose hunting is declining
+## Ontario Moose Hunting is Declining
 
-## Black bears are popular hunting targets in Ontario
+From @fig-huntersmoose, we can see that in all three regions, the number of active hunters has been decreasing since 2006.
+Additionally, the number of moose being harvested has also been decreasing since 2006, which we can see in @fig-harvestmoose.
+This clearly indicates that moose hunting as an activity is declining in all regions of Ontario.
+Although this is happening, we can see from the data that thousands of moose are still being harvested each year.
+However, from the data alone we cannot determine the cause for this decline.
 
-## Weaknesses
+The Ministry of Natural Resources and Forestry reports that moose populations have also been declining since 2004, although the exact reason is unknown [@ctvmoose].
+Some causes of this could be hunting, habitat destruction, disease, or other natural causes.
+Therefore, to protect these populations, the Ministry of Natural Resources and Forestry determines the number of moose hunting tags distributed based on the species population.
+This is likely the main reason for less active moose hunters and harvests over the past 20 years.
+However, despite this decline in moose population, there are still more moose hunters than moose in huntable areas.
+In Ontario, there are around 91,000 moose hunters whereas there are only 78,000 moose in these areas [@moosepop2].
+Despite this decline in moose population and the imbalance between the number of hunters and prey, tags are still being issued.
+In 2023 alone, over 16,000 tags were issued [@moose].
+While it is likely not the sole cause of the declining moose population, hunting activity in Ontario appears to be contributing to this event.
+
+## Black Bears are Popular Hunting Targets in Ontario
+
+In comparison to moose, hunting black bears is not a declining activity.
+However, it appears to have more fluctuation in harvest and active hunting numbers with its peak in 2016.
+On average, more bears are harvested than moose (see @fig-north, @fig-southeast, @fig-southwest) but there are fewer active black bear hunters than moose hunters (see @fig-swhunters, @fig-sehunters, @fig-northhunters).
+Despite this imbalance between hunter and harvest rates, the Ontario black bear population remains high with Ontario having the second-largest black bear population in North America [@bbearon].
+The Ontario black bear population is estimated to be 85,000 to 105,000.
+More black bear sightings have also been reported in southern Ontario near cities and towns [@cbcbbear].
+Black bear hunting in the spring had previously been banned in the 1990s but has reopened with hunters being allowed to hunt as early as May.
+Before the ban, and now that spring hunting has reopened, black bears were popular for trophy hunting [@cbcbbear].
+With high populations and rising spotting of bears near towns and cities, hunting these animals appears to be a method of population regulation.
+Hunters are drawn to them as trophies and it in turn benefits Ontario's economy and helps maintain a healthy black bear population.
+While the data alone does not explain why there are fewer black bear hunters than moose and more black bear harvests than moose harvests, one possible reason could be because those hunting black bears are more successful than moose hunters.
+Additionally, the ratio of hunters to bears is larger than that of hunters to moose.
+
+## Weaknesses and Next Steps
 
@@
 This is because it only focuses on and uses data with two main variables, number of active hunters and harvests.
-From the data alone it is difficult to know the exact reasons for certain trends, such as the decrease in moose harvesting without external research.
+From the data alone, it is difficult to know the exact reasons for certain trends, such as the decrease in moose harvesting without external research.
 This analysis could have been made stronger if paired with moose and black bear population data.
@@
 
-Another weakness is that the data doesn't cover a very large time span.
+Another weakness is that the data does not cover a very large period.
 The black bear data is only from 2012 to 2023 whereas the moose data covers a larger time frame from 2006 to 2023.
 Analyzing a data set with a larger range would provide more depth.
-This would also be beneficial as trends related to populations provide more insight when analyzed over a large period of time.
-The moose data only shows decreasing harvest rates so it would be interesting to see when they began increasing.
+This would also be beneficial as trends related to populations provide more insight when analyzed over a large period.
+The moose data only shows decreasing harvest rates so it would be interesting to see when they began increasing or when their peak was.
 
-## Looking forward
+Moving forward, it is clear that more insights and a deeper analysis could be made with the utilization of more types of data.
+This paper mostly focused on the trends observed in harvest and active hunter data but has created many questions with regard to the causes of these trends.
+These observations are important for Ontario's ecosystem and for maintaining healthy moose and black bear populations.
+Without proper monitoring, these species could be at risk of declining populations or populations that are too large and can in turn impact other species.
+Wildlife hunting is also a big contributor to Ontario's economy and thus healthy populations need to be maintained so that these practices can continue.
 
+\newpage
 # References

### 2024-04-23 scripts/01-data_cleaning.R

@@ -4,7 +4,8 @@
 # Date: 28 March 2024
 # Contact: samantha.barfoot@mail.utoronto.ca
 # License: MIT
-# Pre-requisites: download data from Ontario Data Catalogue
+# Pre-requisites: download data from Ontario Data Catalogue and
+# add to data/raw_data
 
 #### Workspace setup ####
 library(tidyverse)
@@ -132,50 +133,54 @@ all_data_bear <-
   filter(WMU >= 1 & WMU <= 95)
 
 #### SAVE DATA ####
-write_csv(
+
+all_data_bear$WMU <- as.double(all_data_bear$WMU)
+all_data_moose$WMU <- as.double(all_data_moose$WMU)
+
+write_parquet(
   x = north_data_moose,
-  file = "data/analysis_data/north_data_moose.csv"
+  sink = "data/analysis_data/north_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = southeast_data_moose,
-  file = "data/analysis_data/southeast_data_moose.csv"
+  sink = "data/analysis_data/southeast_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = southwest_data_moose,
-  file = "data/analysis_data/southwest_data_moose.csv"
+  sink = "data/analysis_data/southwest_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = all_data_moose,
-  file = "data/analysis_data/all_data_moose.csv"
+  sink = "data/analysis_data/all_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = types_data_moose,
-  file = "data/analysis_data/types_data_moose.csv"
+  sink = "data/analysis_data/types_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = north_data_bear,
-  file = "data/analysis_data/north_data_bear.csv"
+  sink = "data/analysis_data/north_data_bear.parquet"
 )
 
-write_csv(
+write_parquet(
   x = southeast_data_bear,
-  file = "data/analysis_data/southeast_data_bear.csv"
+  sink = "data/analysis_data/southeast_data_bear.parquet"
 )
 
-write_csv(
+write_parquet(
   x = southwest_data_bear,
-  file = "data/analysis_data/southwest_data_bear.csv"
+  sink = "data/analysis_data/southwest_data_bear.parquet"
 )
 
-write_csv(
+write_parquet(
[…]

### 2024-04-23 scripts/01-download_data.R

@@ -1,24 +1,0 @@
-#### Preamble ####
-# Purpose: Downloads and saves the data from [...UPDATE THIS...]
-# Author: Samantha Barfoot
-# Date: 28 March 2024
-# Contact: samantha.barfoot@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: 
-# Any other information needed?
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(stringr)
-
-#### Download data ####
-
-unit <- "75A"
-if(str_ends(unit,"[A-Za-z]")){
-  unit <- str_remove(unit,"[A-Za-z]")
-}
-unit <- strtoi(unit)
-print(unit)
-
-#### Save data ####

### 2024-04-23 scripts/02-test_data.R

@@ -214,7 +214,7 @@ for (i in seq_along(all_bear)){
   }
 }
 
-#TEST 5: Checks that WMU doesn not start with a zero
+#TEST 5: Checks that WMU doesnt not start with a zero
 
 for (i in seq_along(all_moose$WMU)) {
   if (str_starts(all_moose$WMU[i], "0")) {
@@ -229,4 +229,4 @@ for (i in seq_along(all_bear$WMU)) {
 }
 
 #Running Lintr
-#lintr::lint(filename = "scripts/03-test_data.R")
+#lintr::lint(filename = "scripts/02-test_data.R")

### 2024-04-23 scripts/04-model.R

@@ -1,37 +1,0 @@
-#### Preamble ####
-# Purpose: Models... [...UPDATE THIS...]
-# Author: Samantha Barfoot
-# Date: 28 March 2024
-# Contact: samantha.barfoot@mail.utoronto.ca
-# License: MIT
-# Pre-requisites: 
-# Any other information needed?
-
-
-#### Workspace setup ####
-library(tidyverse)
-library(rstanarm)
-
-#### Read data ####
-analysis_data <- read_csv("data/analysis_data/analysis_data.csv")
-
-### Model data ####
-first_model <-
-  stan_glm(
-    formula = flying_time ~ length + width,
-    data = analysis_data,
-    family = gaussian(),
-    prior = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_intercept = normal(location = 0, scale = 2.5, autoscale = TRUE),
-    prior_aux = exponential(rate = 1, autoscale = TRUE),
-    seed = 853
-  )
-
-
-#### Save model ####
-saveRDS(
-  first_model,
-  file = "models/first_model.rds"
-)
-
-

### 2024-04-23 outputs/paper/datasheet.qmd
Finished datasheet and saved data as parquet files

---
title: "Datasheet for 'Wildlife Hunting' data"
author: Samantha Barfoot
thanks: "Code and data are available at: https://github.com/SamanthaBarfoot/wildlife_hunting.git"
date: today
date-format: long
abstract: "This datasheet describes the motivations and composition of the moose and black bear hunting activity and harvest data sets obtained from the Ontario Data Catalogue. It also describes collection process, cleaning process, uses, distribution, and maintenance of the data sets."
format: pdf
number-sections: true
bibliography: references.bib
---

**Motivation**

1. *For what purpose was the data set created? Was there a specific task in mind? Was there a specific gap that needed to be filled? Please provide a description.*
    - The data set was created to enable analysis of Ontario moose and black bear hunting activity. Specifically, the number of active hunters and harvests of both moose and black bears in each Wildlife Management Unit for each year. The data were created to better understand and help manage the moose and black bear population in Ontario with regard to hunting activities. 
2. *Who created the data set and on behalf of which entity?*
    - The creator of the original data set are unknown but they were created on behalf of the Ministry of Resources and Forestry. However, the data sets have been updated and maintained by the Science and Research Branch, specifically by Allen Kwan, Bianca Sayan, Hao Qin, Emily Hill, Eric Giang, Stephanie Whyley, Ryder Leblanc, Cody Schacter, and Razieh Faraji since 2019.
3. *Who funded the creation of the dataset?*
    - The creation of the original data set was funded by the Government of Ontario.

**Composition**

1. *What do the instances that comprise the dataset represent? Are there multiple types of instances?*
	- The instances that comprise the moose data set represent the number of calf, bull, and cow harvests collected in a WMU within a certain year. Additionally, the number of active hunters in that WMU and year is also recorded. The instances that comprise the moose data set represent the number of black bears harvested and the number of active hunters in a WMU within a certain year.
2. *How many instances are there in total?*
	- In total, there are 1229 instances in the moose data set and 1056 instances in the black bear data set.
3. *Does the dataset contain all possible instances or is it a sample (not necessarily random) of instances from a larger set? If the dataset is a sample, then what is the larger set? Is the sample representative of the larger set (for example, geographic coverage)? If so, please describe how this representativeness was validated/verified. If it is not representative of the larger set, please describe why not (for example, to cover a more diverse range of instances, because instances were withheld or unavailable).*
	- TBD
4. *What data does each instance consist of? "Raw" data (for example, unprocessed text or images) or features? In either case, please provide a description.*
	- For moose, each instance consists of the WMU, year, number of active hunters, number of each harvest type (cow, calf, or bull), and the total harvest. For black bears, each instance consists of the WMU, year, number of active hunters, and the total harvest.
5. *Is there a label or target associated with each instance?*
	- There is no label or target associated with each instance.
6. *Is any information missing from individual instances?*
	- There are no values missing from individual instances.
7. *Are there recommended data splits?*
	- There are no recommended data splits for moose or black bear data.
8. *Are there any errors, sources of noise, or redundancies in the data set?*
	- There are no errors, sources of noise, or redundancies in the data set
9. *Is the dataset self-contained, or does it link to or otherwise rely on external resources?*
	- The data sets can be obtained from the Ontario Data Catalogue website at the following links: (https://data.ontario.ca/dataset/moose-hunting-activity-and-harvests) and (https://data.ontario.ca/dataset/black-bear-hunting-activity-and-harvests). These can be downloaded for free by anyone and come with no restrictions and are availble as long as the Goverment of Ontario provides and maintains them.
10. *Does the dataset contain data that might be considered confidential?*
	- The data sets do not contain data that might be considered confidential and can be acessed by the general public. 
11. *Does the dataset contain data that, if viewed directly, might be offensive, insulting, threatening, or might otherwise cause anxiety?*
	- The data sets contain information about animal hunting that certain people might find offensive but the data contains only numbers and years and not detailed information about hunting.
12. *Does the dataset identify any sub-populations?*
	- The moose data set identifies three sub populations, cow, calf, and bull harvests. Each instance contains harvests for each sub-population as well as a total harvest obtained from the sum of the other three harvests. 
13. *Is it possible to identify individuals*
	- No it is not possible to identify individuals as the only data related to people is the number of people, there are no names or other identifying traits used. All the data is anonymous. 
14. *Does the dataset contain data that might be considered sensitive in any way?*
	- Since the data is about hunting and the number of animals harvested each year some individuals may find this offensive due to their personal beliefs about hunting and hurting animals.

**Collection process**

1. *How was the data associated with each instance acquired?*
	- The data was acquired by hunters self reporting the harvests they collected. All licences hunters with tags must do this. Therefore, not all instances can be validated. The number of active hunters is based on how many licences hunters are registered with the Government of Ontario and can thus be validated.
2. *What mechanisms or procedures were used to collect the data?*
	- It is unknown what mechanisms or procedures were used to collect the data. However, since hunters self report the number of harvests it could be assumed that they submitted reports either by mail, phone, or online. 
3. *Who was involved in the data collection process?*
	- It is unknown who exactly was involved in the data collection process, active hunters that reported their harvests remained anonymous as does the original creator of the data sets. 
4. *Over what timeframe was the data collected? Does this timeframe match the creation timeframe of the data associated with the instances?*
	- The data was collected between 2006 and 2023 for moose data and 2012 and 2023 for black bear data. The data sets are also updated each year. These timeframes match the timeframe of the instances as the data.
5. *Did you collect the data from the individuals in question directly, or obtain it via third parties or other sources?*
	- The data was collected from the Ontario Data Catalogue website.
6. *Were the individuals in question notified about the data collection?*
	- They were not notified as the data is freely accessible to the general public.
7. *Did the individuals in question consent to the collection and use of their data?*
	- The data is available for anyone to use and thus they did not need to be notified as it falls under the Open Government Licence.

**Preprocessing/cleaning/labeling**

1. *Was any preprocessing/cleaning/labeling of the data done?*
	- Cleaning of the data was done to group the data into three larger regions: Norhtern, Southeastern, and Southwestern Ontario. Harvest and active hunter data were also combined to find the total across all regions for each year. The WMU values were also cleaned so that they did not contain extra characters such as zeros and letters.
2. *Was the "raw" data saved in addition to the preprocessed/cleaned/labeled data?*
	- The raw data was saved and can be found here: https://github.com/SamanthaBarfoot/wildlife_hunting/tree/main/data/raw_data
3. *Is the software that was used to preprocess/clean/label the data available? If so, please provide a link or other access point.*
	- R studio was used to clean the data, cleaning scripts can be found here: https://github.com/SamanthaBarfoot/wildlife_hunting/tree/main/scripts

**Uses**

1. *Has the dataset been used for any tasks already? If so, please provide a description.*
	- No other tasks that used this data set have been found but some possible tasks that might have used it include other papers and news articles as well as government reports. 
2. *Is there a repository that links to any or all papers or systems that use the dataset? If so, please provide a link or other access point.*
	- No there is not a repository that links all paper or systems that use the data set.
3. *What (other) tasks could the dataset be used for?*
	- The data set could also be used for research papers that analyze bear and moose hunting trends and population trends. Additionally, they could be used to inform news articles and other government reports about hunting activity.
4. *Is there anything about the composition of the dataset or the way it was collected and preprocessed/cleaned/labeled that might impact future uses? Is there anything a dataset consumer could do to mitigate these risks or harms?*
	- All data from the data set is anonymous and only contains counts, no sensitive information. Therefore, there is not anything that might impact future users.
5. *Are there tasks for which the dataset should not be used?*
	- No, as long as the data is not being used to harm someone.

**Distribution**

1. *Will the data set be distributed to third parties outside of the entity on behalf of which the data set was created?*
	- While the data was created by the Ontario Government for monitoring hunting activity, the data was shared publicly online for free and thus third parties can use it.
2. *How will the data set be distributed? Does the dataset have a digital object identifier (DOI)?*
	- The data set was distributed directly on the Ontario Data Catalogue Website as a csv file. There is no DOI
3. *When was the data set distributed?*
	- The data sets were first distributed  31-12-2019.
4. *Will the dataset be distributed under a copyright or other intellectual property (IP) license, and/or under applicable terms of use (ToU)? Provide a link or other access point to, or otherwise reproduce, any relevant licensing terms or ToU, as well as any fees associated with these restrictions.*
	- The data sets are licensed under an Open Government Licence - Ontario (https://www.ontario.ca/page/open-government-licence-ontario)
5. *Have any third parties imposed IP-based or other restrictions on the data associated with the instances?*
	- No third parties imposed IP based or other restrictions on the data associated with the instances
6. *Do any export controls or other regulatory restrictions apply to the data set or to individual instances? *
	- The only restriction is that the source must be acknowledged and you cannot acess personal information (who the hunters are). (https://www.ontario.ca/page/open-government-licence-ontario)

**Maintenance**

1. *Who will be supporting/hosting/maintaining the data set?*
	- The Science and Research Branch of the Ministry of Natural resources and Forestry maintains the data.
2. *How can the owner/curator/manager of the data set be contacted?*
	- They can be contacted online by a form or by email.
3. *Will the data set be updated? How often, by whom, and how updates will be communicated to data set consumers?*
	- The data is updated every year after each hunting season with updates being put on the Open Data Catalogue website. Each update is recorded so users know how up to date the data is.
4. *If the data set relates to people, are there applicable limits on the retention of the data associated with the instances?*
	- The data set does not directly relate to people, only counts of people and thus there are not limits to retention of data as there is no personal information included. 
5. *Will older versions of the data set continue to be supported/hosted/maintained?*
	- Older versions of the data set are not maintained after new updates are released.
6. *If others want to extend/augment/build on/contribute to the dataset, is there a mechanism for them to do so?*
	- There is no mechanism for others to contribute to the data set as it is maintained by the Government of Ontario and having outside parties contribute to it could reduce the reliability of the data.

\newpage

### 2024-04-23 outputs/paper/paper.qmd
Finished datasheet and saved data as parquet files

--- 2024-04-23 paper/paper.qmd
@@
   read_parquet(
-    file = "../data/analysis_data/north_data_bear.parquet"
+    file = "../analysis_data/north_data_bear.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/southeast_data_bear.parquet"
+    file = "../analysis_data/southeast_data_bear.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/southwest_data_bear.parquet"
+    file = "../analysis_data/southwest_data_bear.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/north_data_moose.parquet"
+    file = "../analysis_data/north_data_moose.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/southeast_data_moose.parquet"
+    file = "../analysis_data/southeast_data_moose.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/southwest_data_moose.parquet"
+    file = "../analysis_data/southwest_data_moose.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/all_data_moose.parquet"
+    file = "../analysis_data/all_data_moose.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/types_data_moose.parquet"
+    file = "../analysis_data/types_data_moose.parquet"
   )
@@
   read_parquet(
-    file = "../data/analysis_data/all_data_bear.parquet"
+    file = "../analysis_data/all_data_bear.parquet"
   )

### 2024-04-23 scripts/00-simulate_data.R

@@ -267,49 +267,49 @@ all_data_moose <-
 
 #### SAVE DATA ####
 
-write_csv(
+write_parquet(
   x = north_data_bear,
-  file = "data/analysis_data/north_data_bear.csv"
+  sink = "outputs/analysis_data/north_data_bear.parquet"
 )
 
-write_csv(
+write_parquet(
   x = se_data_bear,
-  file = "data/analysis_data/southeast_data_bear.csv"
+  sink = "outputs/analysis_data/southeast_data_bear.parquet"
 )
 
-write_csv(
+write_parquet(
   x = sw_data_bear,
-  file = "data/analysis_data/southwest_data_bear.csv"
+  sink = "outputs/analysis_data/southwest_data_bear.parquet"
 )
 
-write_csv(
+write_parquet(
   x = north_data_moose,
-  file = "data/analysis_data/north_data_moose.csv"
+  sink = "outputs/analysis_data/north_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = se_data_moose,
-  file = "data/analysis_data/southeast_data_moose.csv"
+  sink = "outputs/analysis_data/southeast_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = sw_data_moose,
-  file = "data/analysis_data/southwest_data_moose.csv"
+  sink = "outputs/analysis_data/southwest_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = types_data_moose,
-  file = "data/analysis_data/types_data_moose.csv"
+  sink = "outputs/analysis_data/types_data_moose.parquet"
 )
 
-write_csv(
+write_parquet(
   x = all_data_bear,
-  file = "data/analysis_data/all_data_bear.csv"
+  sink = "outputs/analysis_data/all_data_bear.parquet"
 )
 
-write_csv(
+write_parquet(
   x = all_data_moose,
-  file = "data/analysis_data/all_data_moose.csv"
+  sink = "outputs/analysis_data/all_data_moose.parquet"
 )
 
 #Running Lintr

### 2024-04-23 scripts/01-data_cleaning.R

@@ -16,10 +16,10 @@ library(arrow)
 library(lintr)
 
 #### Read in data ####
-raw_data_moose <- read_csv("data/raw_data/moose_2023.csv",
+raw_data_moose <- read_csv("inputs/raw_data/moose_2023.csv",
                            show_col_types = FALSE)
 
-raw_data_bear <- read_csv("data/raw_data/black_bear_2023.csv",
+raw_data_bear <- read_csv("inputs/raw_data/black_bear_2023.csv",
                           show_col_types = FALSE)
 
 #### Clean in data ####
@@ -139,47 +139,47 @@ all_data_moose$WMU <- as.double(all_data_moose$WMU)
 
 write_parquet(
   x = north_data_moose,
-  sink = "data/analysis_data/north_data_moose.parquet"
+  sink = "outputs/analysis_data/north_data_moose.parquet"
 )
 
 write_parquet(
   x = southeast_data_moose,
-  sink = "data/analysis_data/southeast_data_moose.parquet"
+  sink = "outputs/analysis_data/southeast_data_moose.parquet"
 )
 
 write_parquet(
   x = southwest_data_moose,
-  sink = "data/analysis_data/southwest_data_moose.parquet"
+  sink = "outputs/analysis_data/southwest_data_moose.parquet"
 )
 
 write_parquet(
   x = all_data_moose,
-  sink = "data/analysis_data/all_data_moose.parquet"
+  sink = "outputs/analysis_data/all_data_moose.parquet"
 )
 
 write_parquet(
   x = types_data_moose,
-  sink = "data/analysis_data/types_data_moose.parquet"
+  sink = "outputs/analysis_data/types_data_moose.parquet"
 )
 
 write_parquet(
   x = north_data_bear,
-  sink = "data/analysis_data/north_data_bear.parquet"
+  sink = "outputs/analysis_data/north_data_bear.parquet"
 )
 
 write_parquet(
   x = southeast_data_bear,
-  sink = "data/analysis_data/southeast_data_bear.parquet"
+  sink = "outputs/analysis_data/southeast_data_bear.parquet"
 )
 
 write_parquet(
   x = southwest_data_bear,
-  sink = "data/analysis_data/southwest_data_bear.parquet"
+  sink = "outputs/analysis_data/southwest_data_bear.parquet"
 )
 
 write_parquet(
   x = all_data_bear,
-  sink = "data/analysis_data/all_data_bear.parquet"
[…]

### 2024-04-23 scripts/02-test_data.R

@@ -16,57 +16,48 @@ library(lintr)
 #### Read in data ####
 
 north_bear <-
-  read_csv(
-    file = "./data/analysis_data/north_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/north_data_bear.parquet"
   )
 
 se_bear <-
-  read_csv(
-    file = "./data/analysis_data/southeast_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/southeast_data_bear.parquet"
   )
 
 sw_bear <-
-  read_csv(
-    file = "./data/analysis_data/southwest_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/southwest_data_bear.parquet"
   )
 
 north_moose <-
-  read_csv(
-    file = "./data/analysis_data/north_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/north_data_moose.parquet"
   )
 
 se_moose <-
-  read_csv(
-    file = "./data/analysis_data/southeast_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/southeast_data_moose.parquet"
   )
 
 sw_moose <-
-  read_csv(
-    file = "./data/analysis_data/southwest_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/southwest_data_moose.parquet"
   )
 
 all_moose <-
-  read_csv(
-    file = "./data/analysis_data/all_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/all_data_moose.parquet"
   )
 
 all_bear <-
-  read_csv(
-    file = "./data/analysis_data/all_data_bear.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/all_data_bear.parquet"
   )
 
 types_moose <-
-  read_csv(
-    file = "./data/analysis_data/types_data_moose.csv",
-    show_col_types = FALSE
+  read_parquet(
+    file = "./outputs/analysis_data/types_data_moose.parquet"
   )
 
 #### Test data ####
